---
title: "Beyond the Ballot: Uncovering the Persistence of Nonvoters in Swiss National Elections"
subtitle: "A Data Analysis non voting"
author: 
  - "Felix Sigrist, felix.sigrist@stud.unilu.ch, 19-641-968"
  - "Supervisor: Dr. Benjamin Schlegel"
lang: en
date: today
date-format: "D. MMMM YYYY"
format: 
  pdf:
    documentclass: scrreprt
    number-sections: true
    toc: true
    toc-depth: 2
    lof: false
    lot: false
    include-in-header:
      - text: |
          \usepackage{authblk}
          \usepackage{titling}
          \RedeclareSectionCommand[beforeskip=5pt plus 1pt minus 1pt]{section}
          \RedeclareSectionCommand[beforeskip=10pt plus 2pt minus 2pt]{chapter}
          \usepackage{pdflscape}
fontsize: 12pt
linestretch: 1.5
geometry: left=2cm,right=2cm,top=2.5cm,bottom=2.5cm
bibliography: nonvoting.bib
citeproc: true
execute:
  echo: false
  warning: false
knitr:
  opts_chunk:
    message: false
---

# Abstract {.unnumbered}

Persistent non-voting threatens democratic legitimacy, yet the drivers of persistent abstention in Switzerland remain underexplored. Drawing on the Civic Voluntarism Model and panel data from the SELECTS survey (2019–2023), six empirical non-voter types are consolidated into three analytically tractable categories: apathetic, protest-oriented, and alternative/satisfied abstainers. A series of multinomial logit models assesses how socio-economic resources (education, income), media consumption, and social and party engagement predict membership in each group. Low formal education strongly increases the odds of apathy-driven abstention; low income is associated with protest-based non-voting; and high media use reduces the likelihood of alternative/satisfied abstention. Regular political discussion and close party proximity consistently lower the probability of persistent non-voting across all categories. These findings highlight structural inequalities and motivational deficits as central barriers to electoral participation and suggest targeted information campaigns and community outreach to re-engage specific non-voter segments.

# Introduction

Although elections are a cornerstone of democratic legitimacy, more than half of the Swiss electorate regularly abstains from voting in federal elections. With turnout consistently below 50 percent since the 1970s (@fig-voter-turnout), abstention has become a structural feature of Swiss democracy [@freitag_labor_2005]. This raises important questions about democratic representation and political inclusion.

```{r}
#| label: "fig-voter-turnout"
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Voter turnout in federal elections has remained below 50% since the early 1970s — illustrating the long-term nature of electoral abstention in Switzerland. Data source: [@noauthor_wahlbeteiligung_nodate]."

library(ggplot2)

# Read in data
Participation <- read.csv("Daten/NRW-Zeitreihe.csv", sep = ";")

# Filter for national-level data
Participation_national <- subset(Participation, geoLevelName == "Schweiz")

# Convert turnout to numeric (handle commas)
Participation_national$wahlbeteiligung <- as.numeric(gsub(",", ".", Participation_national$wahlbeteiligung))

# Plot
ggplot(Participation_national, aes(x = wahl_jahr, y = wahlbeteiligung)) +
  geom_line(color = "#0072B2", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Voter Turnout in Swiss National Council Elections",
    x = "Election Year",
    y = "Turnout (%)"
  ) +
  theme_minimal(base_size = 13) +
  scale_x_continuous(breaks = seq(min(Participation_national$wahl_jahr), max(Participation_national$wahl_jahr), by = 4)) +
  ylim(30, 60)
 

```
Despite its relevance, the phenomenon of non-voting has long been underexplored in Swiss political science. As Bühlmann, Freitag and Vatter [@vatter_pdf_2003] noted two decades ago, Swiss electoral research has shown “astonishing restraint”[@vatter_pdf_2003, 27] in addressing abstention empirically — even though Switzerland, together with the United States, consistently ranks at the bottom of voter turnout among established democracies [@ostrander_using_2021]. @fig-voter-turnout illustrates this trend, showing that turnout in federal elections has remained below the 50 percent mark for over five decades.

Recent research has increasingly challenged the idea that abstention equals apathy. Typological approaches show that non-voters differ substantially in motivation, political attitudes and structural resources [@fatke_wollen_2015]. Studies on selective participation in Switzerland suggest that while some abstain situationally, others remain constantly disengaged [@dermont_taking_2016].

This study focuses on these **persistent non-voters** — individuals who abstained in both the 2019 and 2023 federal elections. It builds on the typology developed by Bühlmann et al. [@vatter_pdf_2003], condensing six original types into three analytically useful categories: **apathetic**, **protest-oriented**, and **satisfied or alternative non-voters**.

Using data from the SELECTS panel (2019–2023) and drawing on the Civic Voluntarism Model [@ostrander_using_2021], this study applies a multinomial logistic regression to explore how political attitudes, social integration and socioeconomic status explain persistent abstention — and how non-voters differ from the voting population and from each other.

Therefore, this paper asks: ***Which socio-economic and socio-political factors predict membership in the three categories of persistent non-voters in the Swiss National Elections of 2019 and 2023?***



# Literature Review

Political non-voting is a central theme in research on Swiss democracy. Several studies have aimed to typologize the heterogeneous group of non-voters and explain their behavior using theoretical models and empirical typologies.

## Classical Approaches to Political Participation

Earlier research on voter turnout was largely grounded in the **socio-structural** approach (e.g., Lazarsfeld), the **social-psychological** model (e.g., Campbell et al.), and **rational choice** theory (e.g., Downs). These were later extended through the **Standard Model of Political Participation** (Barnes & Kaase, 1979), which emphasized the role of socioeconomic resources and cognitive mobilization [@vatter_pdf_2003].

A conceptual advancement came with the introduction of the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which identifies three central factors influencing political participation: **resources**, **motivation**, and **recruitment networks**. The CVM has proven particularly useful in explaining unequal patterns of participation and in identifying different types of abstainers (ebd.).

## Milbrath’s Hierarchical Model of Participation

Milbrath’s framework [@ruedin_testing_2007] conceptualizes political participation as a **cumulative hierarchy**—ranging from passive forms of involvement (e.g., voting, discussing politics) to more active and elite-level activities (e.g., running for office). This model combines individual characteristics with exposure to political stimuli and has been validated through simulation studies that demonstrate the dynamic interplay of individual and institutional factors (ebd.).

## A Swiss Typology: Bühlmann et al. (2003)

[@vatter_pdf_2003] developed one of the first typologies of Swiss non-voters using cluster analysis on survey data from the 1999 federal elections. Their typology includes six distinct types:

-   Politically Disinterested
-   Politically Disenchanted
-   Socially Isolated
-   Politically Incompetent
-   Protest Voters
-   Alternative Participants

These types differ in terms of political interest, efficacy, trust in institutions, and socioeconomic resources.

## Extension and Replication: Fatke & Freitag (2015)

[@fatke_wollen_2015] built upon this typology using SELECTS 2011 data. Their cluster analysis refined the groupings and reaffirmed the heterogeneity of non-voters. Types such as *satisfied but disengaged*, *incompetent*, and *abstaining but participatory* highlight that abstention does not always stem from apathy or alienation, but also from rational disengagement or alternative political expression.

## Democratic Implications and Empirical Consensus

Across all studies, a key insight prevails: **non-voters are not a homogeneous group**. From a democratic theory perspective, the most problematic are the **socially isolated** and **incompetent**, who lack both the resources and social networks for political engagement [@fatke_nichtwahlenden_2015].

At the same time, not all abstention is a cause for concern. "*While there generally is a positive correlation between electoral participation and engagement in other forms of political action, there is an important group of abstainers who chose to participate in other political activities*" [@portos_life_2020, 233]. **Alternative participants** or the **satisfied disengaged**, for example, abstain not due to rejection or apathy, but because they either prefer other forms of participation (e.g., initiatives, petitions) or are fundamentally satisfied with the system.

This differentiated understanding provides the conceptual bridge to the following theory section, which formalizes these empirical distinctions into a testable framework.

------------------------------------------------------------------------

# Theory

## Theoretical framework

### The Civic Voluntarism Model as Theoretical Foundation

This study applies the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which explains political engagement through three core components:

1.  **Resources** – time, money, civic skills, and knowledge\
2.  **Motivation** – interest in politics, efficacy, ideological conviction\
3.  **Recruitment Networks** – social integration, group membership, mobilization

According to CVM, individuals are more likely to participate politically when they are motivated, able, and socially embedded [@ostrander_using_2021]. Conversely, the absence of any one of these factors can lead to abstention [@ostrander_using_2021]. Crucially, access to these dimensions is unequally distributed—meaning that political inequality is reproduced through structural exclusion (ebd.).

The CVM therefore provides a robust foundation for explaining **persistent abstention**, as it integrates individual capacities, perceptions, and social context into one explanatory model.

### Electoral Abstention as a Multi-Dimensional Phenomenon

Political abstention is increasingly seen as a multidimensional phenomenon. Contemporary research emphasizes that non-voters differ systematically in terms of political motivation, cognitive engagement, social integration, and structural access. Typological approaches have shown that abstention reflects distinct profiles of political disengagement, rather than a binary act of participation versus non-participation [@vatter_pdf_2003; @fatke_wollen_2015].

### Persistent Non-Voting as an Analytical Focus

Persistent abstention is of particular interest, as it suggests more entrenched and stable forms of disengagement. Unlike situational abstention, which may result from temporary constraints, persistent non-voters are likely to exhibit deeper structural or attitudinal barriers. Typologizing this group allows for a theory-driven understanding of *why* certain individuals consistently opt out of electoral politics.

## Typologies of Non-Voters in Switzerland

While the Civic Voluntarism Model explains many forms of non-participation through resource or motivation deficits, protest-oriented abstention requires an additional conceptual lens. Drawing on theories of political support [@easton_re-assessment_1975], political scientists argue that abstention may also reflect alienation from democratic institutions. In this view, protest abstention is not a symptom of apathy but rather a passive expression of dissatisfaction — a silent refusal to endorse a political system perceived as unresponsive, unfair, or corrupt (ebd.).

This distinction is particularly relevant in the Swiss context, where institutional trust is generally high, but perceived political inefficacy may still motivate strategic disengagement. Incorporating this dimension allows for a clearer differentiation between those who are disengaged and those who abstain deliberately out of protest.

Furthermore, the group of those who participate in alternative political ways and / or are satisfied with the system can be explained by the "low-salience" theory which highlights the attractiveness of alternative forms of political participation in Switzerland [@vatter_pdf_2003, 49].In the special case of Switzerland, this group may also regularly participate at local elections / votations, is satisfied with the system and esteem their political impact as high (ebd.)

Building on the CVM, theories of political support, low-salience theory and empiric findings, this section presents the typology of persistens non-voters which serves as the dependent variable in this paper. The typologies developed by Bühlmann et al. [@vatter_pdf_2003] and extended by Fatke & Freitag [@fatke_wollen_2015] offer a detailed framework for classifying non-voters. For the purpose of this study, these types are aggregated into three theoretically meaningful categories:

| Type | Main Mechanism | Theoretical Foundation |
|------------------------|------------------------|------------------------|
| **Apathetic** | Lack of motivation/resources | Civic Voluntarism Model |
| **Protest-oriented** | Political alienation | System support/trust theory |
| **Alternative/Satisfied** | Rational disengagement | low salience theory |

This typology provides the conceptual basis for analyzing **persistent non-voters**, i.e., those who abstained in both 2019 and 2023 federal elections in Switzerland.

## Theoretical Expectations and Hypotheses

In order to explain the differences between the three types of abstention, testable hypotheses are derived from the Civic Voluntarism Model and the typology shown above. All hypotheses will be tested via a multinomial logistic regression, with **voters** as the reference category.

-   **H1 (Education & Resources):** Individuals with lower formal education are more likely to be *apathetic non-voters* than voters or other non-voter types.

-   **H2 (Income & Political Alienation):** Individuals with lower household income are more likely to be *protest-oriented non-voters* than voters or other non-voter types.

-   **H3 (Media Use & Low Salience):** Higher levels of media consumption significantly reduce the likelihood of being an *alternative/satisfied non-voter*, but have little effect on apathy- or protest-based non-voting.

-   **H4 (Political Discussion & Apathy vs. Protest):** Frequent political discussion decreases the probability of *apathy-based non-voting* more strongly than it decreases the probability of *protest-oriented non-voting*.

-   **H5 (Party Membership & Integration):** Party members are significantly less likely to be *apathetic* or *protest-oriented non-voters* compared to voters.

These hypotheses will be empirically tested using a multinomial logistic regression model, treating **voters** as the reference category. The predictors reflect the theoretical dimensions discussed above.

# Empirical Analysis

## Data and Descriptive Statistics

The analysis is based on data from the Swiss Election Study (SELECTS) panel survey, covering the 2019 and 2023 federal elections. The dataset includes respondents who participated in all relevant waves and provides detailed information on political attitudes, behavior, and sociodemographic background.

To identify persistent abstainers, respondents who reported not voting in both the 2019 and 2023 National Council elections were selected. Voting behavior in 2019 is captured by the variable `w1_f10200rec` (re-coded: 1 = voted, 0 = did not vote), and in 2023 by `w3_f11100`. Respondents who reported voting in both elections were classified as **voters** and serve as the reference category in the analysis.

## Operationalization of Persistent Non-Voter Types

To test our typology of persistent non-voters, we assign each respondent—who abstained in both the 2019 and 2023 federal elections—to one of three theoretically derived groups, or to the **voter** reference category. Assignment is based on “scores” computed from sets of survey items that best capture each type’s defining mechanism. Cases without a clear unique maximum score are excluded.

### Apathetic Non-Voters

Grounded in the Civic Voluntarism Model, this type reflects a deficit in motivation or resources. We operationalize apathy as the count of affirmative responses to items indicating:\
- **Procedural complexity** (“Elections too complicated”)\
- **General lack of interest** (“Not interested in politics”)\
- **Information gap** (“Did not know candidates”)\
- **Low campaign engagement** (“Paid little attention to the campaign”)\
- **Self-reported disinterest** (“Rather not interested in politics”)

Respondents whose apathy score exceeds their scores on all other dimensions are classified as **apathetic non-voters**.

### Protest-Oriented Non-Voters

Building on theories of political support and democratic legitimacy [@easton_re-assessment_1975], protest abstention captures passive expressions of system alienation. We measure this alienation through indicators of:\
- **Party/system distrust** (“Not convinced by any party/candidate”)\
- **Perceived inefficacy** (“Elections do not change anything”)\
- **Low institutional trust** (federal council, assembly, and national parties rated in the bottom trust categories)\
- **Dissatisfaction with democracy** (low overall satisfaction)

Those with the highest protest-oriented score are assigned to the **protest** group.

### Alternative/Satisfied Non-Voters

Drawing on low-salience theory, this category encompasses individuals who are content with the political system or engage through non-electoral channels [@vatter_pdf_2003]. We capture this “satisfied disengagement” via:\
- **Alternative participation** (petition, initiative or referendum signatures)\
- **Self-reported interest** (“Interested in politics”)\
- **High satisfaction with democracy**\
- **Membership in civic or political organizations** (unions, environmental, human rights, consumer, employers’, party activities)\
- **High trust in national parties**

Respondents whose alternative/satisfied score dominates their other scores form the **satisfied** group.

### Voters (Reference Category)

All respondents who report voting in both 2019 and 2023 comprise the **voter** category and serve as our reference group in subsequent multinomial logistic regressions.

This operationalization ensures a clear alignment between our theoretical typology—CVM for apathy, system-support theory for protest, and low-salience for alternative engagement—and the empirical classification of persistent non-voters.

Respondents who did not fall clearly into one of these categories were excluded from the analysis. The resulting dependent variable `vote_group` consists of four categories: `voter`, `apathetic`, `protest`, and `satisfied`.

Table 4.1 below summarizes the group sizes:

```{r}
#| echo: false
#| results: false
library(tidyverse)
library(haven)
library(janitor)
library(naniar)

# Load and clean dataset
df <- read_sav("Daten/2626_Selects2023_Panel_Data_v2.0.sav") %>%
  zap_labels() %>%           
  clean_names()              

# Identify persistent non-voters (did not vote in both 2019 and 2023)
df_nv <- df %>%
  filter(w1_f10200rec == 0, w2_3_voted == 0) #W2_3_voted #w3_f11100

library(dplyr)

df_nv <- df_nv %>%
  mutate(
    # Scores berechnen (Anzahl zutreffender Gründe pro Typ)
    score_protest = (w3_f11304 == 1) + #Reason non voting: Not convinced by any party/ candidate
                    (w3_f12800a %in% c(0,4)) + #trust in federal council (1-10)
                    (w3_f12800b %in% c(0,4)) + #Trust: Federal Assembly (1-10)
                    (w3_f11309 == 1)+ #Reason non voting: Elections do not change anything
                    (w3_f12800c %in% c(0,4))+ #Trust: Political parties at national level
                    (w1_f13700 %in% c(0,2)) , #Satisfaction with democracy

    score_apathetic = 
                      (w3_f11303 == 1)+#Reason nonvoting: Elections too complicated
                      (w3_f11302  == 1)+#Not interested in politics
                      (w3_f11306  == 1)+#did not know candidates
                      (w2_f10160  == 1)+#attention to election campaign
                      (w3_f10100  %in% c(1,2)) #rather not interested in politics
                      , #Union or employee nun organization: Member

    score_satisfied = (w3_f12610a == 1) + #Signed a petition, popular initiative or a referendum (past 5 years)
                      (w3_f11302  == 2) + #interested in politics
                      (w1_f13700 %in% c(3,4)) , #Satisfaction with democracy
                      (w3_n13_030g2 == 1) + #Political party: participated in activities
                      (w3_n13_030e1 == 1) + #Environmental organization: member
                      (w3_n13_030d1 == 1) + #Human rights or peace organization: member
                      (w3_n13_030c1 == 1) + #Consumer organization: member
                      (w3_n13_030b1 == 1) + #Economic, employers' or farmers' association: member
                      (w3_n13_030a1 == 1) + #Union or employee organization: Member
                      (w3_f10100  %in% c(3,4)) + #R's interest in politics, rather interested
      #R's interests in politics
                      (w3_f12800c %in% c(7,10)), #Trust: Political parties at national level

    # finale Gruppenzuordnung nach Maximal-Score
    vote_group = case_when(
      score_protest   == pmax(score_protest, score_apathetic, score_satisfied) ~ "protest",
      score_apathetic == pmax(score_protest, score_apathetic, score_satisfied) ~ "apathetic",
      score_satisfied == pmax(score_protest, score_apathetic, score_satisfied) ~ "satisfied",
      TRUE ~ NA_character_
    )
  )


# Add consistent voters
df <- df %>%
  mutate(
    vote_group = case_when(
      w1_f10200rec == 1 & w3_f11100 == 4 ~ "voter",
      TRUE ~ NA_character_
    )
  )

# Merge voter types back into main dataset
df$vote_group[df$id %in% df_nv$id] <- df_nv$vote_group

# Create all additional variables BEFORE filtering
# Create all additional variables BEFORE filtering
df <- df %>%
  mutate(
    # Z-standardized predictors
    interest_z = scale(w3_f10100)[,1],
    trust_council_z = scale(w3_f12800a)[,1],    # Federal Council
    trust_parliament_z = scale(w3_f12800b)[,1], # Parliament
    discussion_z = scale(w2_f13310)[,1],        # from Wave 2

    # Media use index from 5 sources (Wave 3)
    media_index = rowMeans(select(., w3_f12610a, w3_f12610b, w3_f12610d, w3_f12610g, w3_f12610i), na.rm = TRUE),
    media_index_z = scale(media_index)[,1],

     # Dummy: party membership (1 = yes if code 10–350, else 0)
    membership_dummy = if_else(
      between(w3_f14081rec, 10, 350),
      1L,
      0L,
      missing = 0L
    ),

    # Categorical predictors
    vote_group = relevel(factor(vote_group), ref = "voter"),
    education = w1_f21310rec,
    income = w1_f28910,
    employment = w1_f21400,

    # Grouped categories
    edu_group = case_when(
      education %in% c(1, 2, 3,4) ~ "low",
      education %in% c(5, 6,7,8,9) ~ "mid",
      education %in% c(10, 11, 12) ~ "high",
      TRUE ~ "unknown"
    ),
    income_cat = case_when(
      income %in% c(1, 2, 3, 4) ~ "low",
      income %in% c(5, 6, 7) ~ "mid",
      income %in% c(8, 9, 10, 11,12,13,14,15,16,17,18,19,20) ~ "high",
      TRUE ~ "unknown"
    ),
    job_cat = case_when(
      employment == 1 ~ "fulltime",
      employment == 2 ~ "parttime",
      employment %in% c(3, 4, 5, 6, 7, 8) ~ "other",
      TRUE ~ "unknown"
    )
  )


# Final filtered dataset for modelling
df_model <- df %>%
  filter(!is.na(vote_group))

df_model <- df_model %>%
  filter(
    complete.cases(
      interest_z,
      trust_parliament_z,
      trust_council_z,
      edu_group,
      income_cat,
      job_cat,
      media_index_z,
      membership_dummy,
      discussion_z,
      weight_tot,
      w1_f14000
    )
  )


# Show group distribution
table(df_model$vote_group, useNA = "always")


```

```{r}
#| label: "table-votegroups"
#| echo: false
#| results: asis
#| warning: false
#| message: false

library(dplyr)
library(knitr)

df_model %>%
  count(vote_group, name = "N") %>%
  arrange(desc(N)) %>%
  mutate(share = round(100 * N / sum(N), 1)) %>%
  kable(
    caption = "Distribution of vote groups in the analytical sample",
    col.names = c("Group", "N", "Share (%)")
  )

```

The distribution confirms that **apathetic non-voters** represent the largest group in the analytical sample, followed by **satisfied / alternative** and **protest** non-voters. As a reference group, all consistent voters (whom participated in 2019 and 2023) were added. This distribution reflects the empirical prevalence of non-voting types among persistent abstainers and underlines the heterogeneity of non-voter profiles. This also suggests that motivational and resource-related non-voting could be more widespread than ideologically motivated protest or strategic satisfaction.

## Methods

To test the proposed hypotheses, three multinomial logistic regression models are estimated. This modeling approach is appropriate for nominal dependent variables with more than two categories. The dependent variable in this analysis is `vote_group`, which consists of the four categories that are theoretically grounded were explained in the previous section: **voters** (reference category), **apathetic**, **protest-oriented**, and **satisfied/alternative** non-voters.

The general form of the multinomial logistic regression model is:

$$
\log\left( \frac{P(Y = k)}{P(Y = \text{Voter})} \right) = \beta_{0k} + \beta_{1k}X_1 + \beta_{2k}X_2 + \dots + \beta_{pk}X_p, \quad \text{for } k \in \{\text{Apathetic, Protest, satisfied}\}
$$

Here, (Y) denotes the categorical outcome variable (`vote_group`), and the coefficients (\beta\_{ik}) represent the effect of predictor (X_i) on the log-odds of being in category (k) compared to the voter reference category.

The analysis is conducted in three steps, each corresponding to a model with increasing complexity:

-   **Model 1** includes key socioeconomic indicators: educational level, income and occupational status.
-   **Model 2** adds indicators related to the media use in regards to political communication: media useage.
-   **Model 3** adds variables related to political integration and communication: party membership, frequency of political discussions and closeness to a party.

All models are estimated using **multinomial logistic regression** with robust standard errors. The interpretation of coefficients is based on **log-odds** relative to the reference category (**voters**). For ease of interpretation, average marginal effects (AMEs) are reported in the results section.

Prior to analysis, all continuous predictors are standardized (mean = 0, sd = 1). Categorical predictors are converted to factors. Missing values are handled using listwise deletion. The final analytical sample consists of 589 cases with a valid `vote_group` classification and complete data on the relevant predictors.

## Variable Operationalization

The dependent variable `vote_group` is a categorical variable with four values: **voter** (reference category), **apathetic**, **protest-oriented**, and **satisfied/alternative** non-voters. It is constructed based on self-reported turnout behavior in 2019 (`w1_f10200rec`) and 2023 (`w3_f11100`), as well as respondents’ stated reasons for abstaining in 2023 (`w3_f11301` to `w3_f11310`).

The independent variables are grouped according to the logic of the Civic Voluntarism Model and introduced step-by-step in three models:

### Model 1: Socioeconomic baseline

The first model introduces structural indicators of socioeconomic status, representing the “resources” dimension of the Civic Voluntarism Model. These variables are operationalized as follows:

-   **Education Level** (`w3_f21310`): Highest level of completed education, recoded to an ordinal scale (from compulsory schooling to tertiary education).
-   **Income** (`w3_f22000_rec`): Monthly household income, recoded into 11 ordered categories.
-   **Employment Status** (`w3_f21000`): Employment situation (e.g., full-time, unemployed, student), recoded into a categorical factor.

All predictors are entered as ordinal or categorical variables. This model tests whether socioeconomic disadvantages are associated with persistent abstention.

### Model 2: Media usage extension

-   **Media Use** (`media_index_z`): Aggregated and standardized frequency of political news consumption via various media (newspapers, TV, internet)

### Model 3: Political Integration Extension

-   **Political Discussion Frequency** (`discussion_z`): Standardized measure of how frequently respondents discuss politics with friends or family (5-point scale).
-   **Party membership** (`membership_dummy`): A dummy variable created out of `w3_f14081rec` which indicates whether someone is a member of a political party or not. -**Party closeness** (`w1_f14000`): A dummy variable that indicates if someone feels close to a political party or not.

## Results

In this section, the specification and estimation of three nested multinomial logistic regression models—distinguishing actual voters, apathetic non-voters, protest non-voters and satisfied non-voters—are presented. Model comparisons are conducted using Akaike Information Criterion (AIC), Bayesian Information Criterion (BIC) and McFadden’s pseudo-R², and improvements in fit are evaluated via likelihood-ratio tests. Thereafter, the estimated log-odds coefficients for key predictor blocks (socioeconomic indicators, media usage and political integration measures) are interpreted to elucidate how each set of variables incrementally contributes to an understanding of non-voting behavior.

```{r include=FALSE, trace=FALSE, message=FALSE, warning=FALSE}
library(nnet)

m1 <- multinom(
  vote_group ~ edu_group + income_cat + job_cat,
  data    = df_model,
  #weights = w4_longweight,
  trace   = FALSE,
  na.action = na.exclude
)

m2 <- multinom(
  vote_group ~ edu_group + income_cat + job_cat+
    media_index_z,
  data    = df_model,
  #weights = w4_longweight,
  trace   = FALSE,
  na.action = na.exclude
)




# --- Model 3 neu schätzen ------------------------------------------------------
m3 <- multinom(
  vote_group ~ edu_group + income_cat + job_cat+
    media_index_z + membership_dummy + discussion_z+w1_f14000,
  data    = df_model,
  #weights = w4_longweight,
  trace   = FALSE,
  na.action = na.exclude
)

invisible(NULL)      # letzte Zeile gibt NULL zurück → garantiert kein Output






```

```{=latex}
\begin{landscape}
```

```{r}


library(modelsummary)
library(kableExtra)

coef_map <- c(
  # intercepts
  "(Intercept)" = "Intercept",
  #other
  "interest_z"         = "Political interest (z)",
  "trust_parliament_z" = "Trust in parliament (z)",
  "trust_council_z"    = "Trust in fed. council (z)",
  "edu_groupmid"       = "Middle education (vs. high)",
  "edu_grouplow"       = "Low education (vs. high)",
  "income_catmid"      = "Middle income (vs. high)",
  "income_catlow"      = "Low income (vs. high)",
  "job_catparttime"    = "Part-time work (vs. fulltime)",
  "job_catother"       = "Other work (vs. fulltime)",
  "media_index_z"      = "Media use index (z)",
  "membership_dummy"   = "Party membership (yes)",
  "discussion_z"       = "Political discussion (z)",
  "w1_f14000" = "Closeness to a party (yes)"
)


tbl <- modelsummary(
  list(
    "Model 1 (Socioeconomic base)"            = m1,
    "Model 2 (+Media usage)" = m2,
    "Model 3 (+Political Integration)"= m3
  ),
  shape    = term ~ model + response,
  coef_map = coef_map,
  estimate = "{estimate}{stars}",
  statistic= "({std.error})",
  fmt      = 2,
  gof_map  = c("nobs" = "N", "AIC" = "AIC", "BIC" = "BIC"),
  gof_omit = "Deviance",
  title    = "Multinomial logit of persistent non-voting (ref. = voters)",
  output   = "kableExtra"
)


tbl |>
  footnote(
    general = c(
      "Coefficients are log-odds relative to voters (reference category).",
      "Standard errors in parentheses.",
      "Positive coefficients = higher likelihood of belonging to the non-voter type."
    ),
    symbol            = c("*** p < .001", "** p < .01", "* p < .05"),
    threeparttable    = TRUE,
    footnote_as_chunk = TRUE,
    escape            = FALSE
  ) |>
  kable_styling(latex_options = "hold_position")

```

```{=latex}
\end{landscape}
```

Table 4.2 shows the results from the three multinomial logistic regression models in which the three types of persistent non-voters are each compared to the reference category of persistent voters. This table was created using R packages lmtest, dplyr, tibble and knitr [^1] .

[^1]: @Zeileis2002, @Wickham2023dplyr, @Mueller2023tibble, @Xie2024knitr

The results displayed in this table can be summarized as following:

### Model 1 (Socioeconomic baseline)

Individuals with low education have substantially higher log-odds of non-voting than those with high-education. This is especially the case for the persistent non-voter types apathetic ( $\beta = 1.08$, *p* \< .001 ) and satisfied ( $\beta = 1.17$, *p* \< .05 ). Middle education shows a smaller and generally non-significant effect. Low income significantly increases the likelihood of all three non-voter types ( $\beta$ between 0.95 and 1.72, *p* \< .01 ). Middle income only reaches significance for the protest type. Both part-time workers and those in "other work" are significantly less likely to be persistent non-voters (e.g. for apathetic $\beta = -0.73$ and $\beta = -1.45$, *p* \< .001 ).

### Model 2 (+Media use)

The standardized media-use index has a strong negative effect on the probability of beloning to the "satisfied" non-voter group ( $\beta = -1.87$, *p* \< .001 ), which means that higher media engagement makes one less likely to be associated with the satisfied non-voter type. The coefficients for the apathetic and protest non-voter types remain small and not statistically significant.

### Model 3 (+Political Integration)

Individuals who feel close to a party have a lowered likelihood of being associated with any of the persistent non-voting types. More frequent political discussion strongly reduces the odds of being associated with any type of non-voter compared to belonging to the persistent voter group. The estimates for party membership show signs of perfect separation. This is the case as none of the persistent non-voters are member in a political party, except for one satisfied-non voter. These estimates should therefore be interpreted with caution.

```{r model-comparison-tables, echo=FALSE, message=FALSE}


library(tibble)
library(lmtest)
library(knitr)
library(dplyr)

# -- Fit table --------------------------------------------------------
fit <- tibble(
  Model       = c("Model 1", "Model 2", "Model 3"),
  N           = c(nrow(model.frame(m1)), nrow(model.frame(m2)), nrow(model.frame(m3))),
  AIC         = c(AIC(m1), AIC(m2), AIC(m3)),
  BIC         = c(BIC(m1), BIC(m2), BIC(m3)),
  McFadden_R2 = c(
    round(1 - as.numeric(logLik(m1)) / as.numeric(logLik(update(m1, .~1))), 3),
    round(1 - as.numeric(logLik(m2)) / as.numeric(logLik(update(m2, .~1))), 3),
    round(1 - as.numeric(logLik(m3)) / as.numeric(logLik(update(m3, .~1))), 3)
  ),
  Df = c(attr(logLik(m1), "df"), attr(logLik(m2), "df"), attr(logLik(m3), "df"))
)

kable(
  fit,
  caption  = "Model fit statistics (multinomial logit)",
  format   = "latex",
  booktabs = TRUE,
  digits   = c(0, 0, 2, 2, 3, 0)
)
```

```{r echo=FALSE, message=FALSE}


# -- Likelihood‐Ratio Tests combined ---------------------------------
lr1 <- lrtest(m1, m2)
lr2 <- lrtest(m1, m3)
lr3 <- lrtest(m2, m3)

lr_all <- bind_rows(
  tibble(
    Comparison   = "Model 1 vs. Model 2",
    ΔDf          = m2$edf - m1$edf,
    Chisq        = round(lr1$Chisq[2], 2),
    `Pr(>Chisq)` = sprintf("%.2e", lr1$`Pr(>Chisq)`[2])
  ),
  tibble(
    Comparison   = "Model 1 vs. Model 3",
    ΔDf          = m3$edf - m1$edf,
    Chisq        = round(lr2$Chisq[2], 2),
    `Pr(>Chisq)` = sprintf("%.2e", lr2$`Pr(>Chisq)`[2])
  ),
  tibble(
    Comparison   = "Model 2 vs. Model 3",
    ΔDf          = m3$edf - m2$edf,
    Chisq        = round(lr3$Chisq[2], 2),
    `Pr(>Chisq)` = sprintf("%.2e", lr3$`Pr(>Chisq)`[2])
  )
)

kable(
  lr_all,
  caption   = "Likelihood‐Ratio Tests for All Model Comparisons",,
  format    = "latex",
  booktabs  = TRUE,
  digits    = c(0, 0, 2, 3)
)

```

Table 4.3 reports the fit statistics for the three progressively specified multinomial logit models predicting the persistent non-voter categories, with persistent voters as the reference group. Model 1, which includes only socioeconomic predictors, yields an Akaike Information Criterion (AIC) of 2799.45, a Bayesian information Criterion (BIC) of 2989.88 and a McFadden's pseudo-R² of 0.07 (30 degrees of freedom). Upon adding a standardized media-use index in Model 2, the AIC falls to 2747.52, the BIC to 2957.00 and McFadden’s R² rises to 0.089 (33 degrees of freedom), indicating that media consumption explains additional variance in non-voting behavior. The inclusion of political-integration measures (party membership, political discussion and party closeness) in Model 3 yields a further decline in AIC to 2336.93 and BIC to 2603.54, while McFadden’s R² more than doubles to 0.235 (42 degrees of freedom). These systematic reductions in AIC and BIC, together with the substantial increase in pseudo-R², demonstrate that each block of variables—first media usage, then political integration—meaningfully enhances the model’s capacity to distinguish between different non-voter types and voters, despite the growing number of parameters.

This is also confirmed by Table 4.4 which presents likelihood-ratio tests for pairwise comparisons among the three models. Comparing Model 1 against Model 2 (ΔDf = 3) yields χ² = 57.92 (p = 1.63 × 10⁻¹²), confirming that the addition of media usage significantly improves model fit over the socioeconomic baseline. The test of Model 1 versus Model 3 (ΔDf = 12) produces χ² = 486.52 (p = 1.64 × 10⁻⁹⁶), and the comparison of Model 2 versus Model 3 (ΔDf = 9) gives χ² = 428.59 (p = 1.08 × 10⁻⁸⁶). All p-values are far below conventional significance thresholds, indicating that political-integration variables contribute a highly significant improvement in explaining non-voting behavior beyond both socioeconomic characteristics and media usage. Consequently, Model 3 represents the best-fitting specification among the three.

```{r}
#| label: tbl-hypotheses-clean
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(knitr)

# Hypotheses table with updated support statuses
hypotheses_clean <- tibble::tibble(
  Hypothesis  = c("H1", "H2", "H3", "H4", "H5"),
  Description = c(
    "Low education increases likelihood of apathetic non-voting",
    "Low income increases likelihood of protest-oriented non-voting",
    "High media use reduces likelihood of alternative/satisfied non-voting",
    "Political discussion reduces apathy more strongly than protest",
    "Party membership reduces apathetic and protest non-voting"
  ),
  Supported = c(
    "Supported (* p<.05)",
    "Supported (*** p<.001)",
    "Supported (*** p<.001)",
    "Supported (*** p<.001 for apathy; ** p<.01 for protest)",
    "Supported with caution (perfect separation; caution: protest N = 47)"
  )
)

kable(
  hypotheses_clean,
  caption   = "Summary of Hypotheses and Empirical Support (Model 3)",
  col.names = c("Hypothesis", "Description", "Empirical Support"),
  align     = c("c", "c", "c")
)


```

In summary, **Model 3 delivers the most precise and theory-aligned estimates**, confirming all five hypotheses: lower education predicts apathetic non-voting (+0.85*, p \< .05), low income drives protest abstention (+1.52***, p \< .001), high media use sharply reduces alternative/satisfied non-voting (–1.66***, p \< .001), political discussion curtails apathy more than protest (–2.22*\*\* vs. –0.74**, p \< .001/0.01), and party membership perfectly separates out apathetic and protest voters (perfect separation in N=226/47). While** Models 1 and 2\*\* already demonstrated the core motivational and socioeconomic mechanisms, **Model 3** not only improves fit but also **quantifies** each theoretical expectation—thereby reinforcing and deepening our understanding rather than altering the substantive conclusions.

# Discussion
Several central insights emerge from the analyses. First, resource-related factors continue to shape distinct non-voter profiles in line with the Civic Voluntarism Model. Individuals lacking formal education tend to withdraw from electoral contests out of disengagement, whereas those facing financial constraints are more likely to abstain as an expression of protest. These patterns suggest that both a shortage of personal resources and underlying political discontent serve as enduring barriers to turnout.

Second, measures of media engagement add nuance to this picture. Those who frequently consume political news and commentary are noticeably less inclined to abstain simply because they feel content with the status quo. In other words, exposure to a variety of information sources appears to channel potential abstainers toward alternative forms of civic involvement—such as online discussions or community initiatives—rather than non-participation at the ballot box.

Third, social-motivational variables underscore the protective role of political embeddedness. Regular discussion of public affairs and active ties to political organizations both work to diminish apathy-driven withdrawal and protest-oriented non-voting. Remarkably, very few abstainers report strong party affiliations, highlighting the importance of interpersonal recruitment networks in sustaining electoral engagement.

Taken together, model-fit comparisons confirm that combining resource, media, and social-integration predictors significantly improves the ability to distinguish among different non-voter types. These findings indicate that structural constraints, informational exposure, and social ties operate in concert to produce persistent patterns of abstention, rather than acting independently.

From a theoretical perspective, the results extend existing turnout models by integrating ideas from system-trust research and low-salience theory. This synthesis yields a more differentiated understanding of who abstains and why, challenging one-dimensional approaches. On a practical level, the evidence points toward tailored interventions: low-threshold informational tools—such as clear, jargon-free voting guides for less-educated audiences—and community forums for local deliberation may re-engage apathetic or contented abstainers. Meanwhile, transparent feedback mechanisms that illustrate how individual votes translate into policy outcomes could address the frustrations of protest-driven non-voters.

Limitations of the study include exclusive reliance on self-reported data and the reduction of complex motivational forces into composite indices, which may obscure subtle psychological processes. Future research should therefore adopt mixed-methods designs, combining in-depth interviews to explore emotional and cognitive drivers with field experiments to assess which mobilization strategies work best under varying conditions. By doing so, scholars can refine both theory and practice, paving the way for more inclusive democratic participation.


# Conclusion
This study deepens understanding of electoral abstention by augmenting the Civic Voluntarism Model with insights from media engagement and social embeddedness. Its chief innovation lies in weaving together resource-based explanations with perspectives on system trust and issue salience, thereby moving beyond a one-dimensional account of who abstains to illuminate why different segments opt out of voting.

The analysis reveals that deficits in personal resources and levels of formal education shape distinct forms of non-voting. Lower educational attainment tends to foster disengagement and apathy, whereas constrained economic circumstances drive abstention motivated by dissatisfaction or protest. Introducing measures of media engagement highlights the power of information exposure to counteract contentment-based non-participation: those who routinely consume political news and commentary are drawn toward alternative civic activities rather than withdrawing from the ballot. Equally significant is the role of social recruitment networks. Individuals embedded in political conversations and formal party structures exhibit markedly lower tendencies to abstain, whether through apathy or protest, underscoring how interpersonal ties and organizational affiliation function as catalysts for turnout.

In answering the central research question—**Which socio-economic and socio-political factors explain membership in the three categories of persistent non-voters?**—the models show that low education most strongly predicts apathy-driven abstention, low income corresponds to protest-driven abstention, and strong party ties plus regular political discussion consistently reduce the likelihood of any form of persistent non-voting. Media engagement further distinguishes the satisfied/alternative group by linking high news consumption to continued civic involvement outside the ballot box. These insights confirm that resource constraints and embeddedness jointly shape who abstains and why.

These findings carry important theoretical and practical implications. By demonstrating that resources, informational access, and social-motivational factors operate in concert, the study extends the reach of the Civic Voluntarism Model into a more nuanced framework that captures the heterogeneous nature of non-voters. From a policy perspective, the results suggest tailored interventions—concise voting guides for the less educated, deliberative forums for the apathetic, transparent feedback for protest-oriented abstainers, and integrative outreach to connect alternative participants back to the ballot.

Despite these advances, the study’s reliance on self-reports and composite indices presents limitations, and its cross-sectional design constrains causal inference. Future mixed-method research—combining qualitative interviews with field experiments—can unpack the cognitive and emotional drivers of each non-voter type and rigorously test these mobilization strategies, advancing efforts toward a more inclusive and vibrant democracy.


\`\`

\newpage

# References

::: {#refs}
:::

# Appendix

## Average marginal effects

```{r ame-longtable, echo=FALSE, results='asis', message=FALSE}
library(dplyr)
library(kableExtra)
library(knitr)

# 1) Daten laden und vorbereiten -----------------------------------------
ame_m1 <- readRDS("Daten/ame_m1.rds") |> mutate(model = "Model 1")
ame_m2 <- readRDS("Daten/ame_m2.rds") |> mutate(model = "Model 2")
ame_m3 <- readRDS("Daten/ame_m3.rds") |> mutate(model = "Model 3")

ame_all <- bind_rows(ame_m1, ame_m2, ame_m3) %>%
  filter(!is.na(group) & group != "voter") %>%
  mutate(
    AME = estimate * 100,
    SE  = std.error * 100
  )

# 2) pred_key & Predictor erzeugen --------------------------------------
coef_map <- c(
  "interest_z"           = "Political interest (z)",
  "trust_parliament_z"   = "Trust in parliament (z)",
  "trust_council_z"      = "Trust in federal council (z)",
  "edu_grouplow"         = "Compulsory education (vs. high)",
  "edu_groupmid"         = "Secondary education (vs. high)",
  "edu_groupunknown"     = "Unknown education level (vs. high)",
  "income_catlow"        = "Low household income (vs. high)",
  "income_catmid"        = "Middle household income (vs. high)",
  "income_catunknown"    = "Unknown household income (vs. high)",
  "job_catother"         = "Other employment status (vs. full-time)",
  "job_catparttime"      = "Part-time employment (vs. full-time)",
  "job_catunknown"       = "Unknown employment status (vs. full-time)",
  "media_index_z"        = "Media use index (z)",
  "membership_dummy"     = "Party membership (yes vs. no)",
  "discussion_z"         = "Political discussion (z)"
)

tbl_long <- ame_all %>%
  mutate(
    pred_key = case_when(
      contrast %in% c("low - high", "mid - high", "unknown - high") ~
        paste0(term, sub(" -.*$", "", contrast)),
      contrast %in% c("other - fulltime", "parttime - fulltime", "unknown - fulltime") ~
        paste0(term, sub(" -.*$", "", contrast)),
      TRUE ~ term
    ),
    Predictor = recode(pred_key, !!!coef_map, .default = pred_key),
    AME       = sprintf("%.2f", AME),
    SE        = sprintf("%.2f", SE)
  ) %>%
  select(Model = model, Outcome = group, Predictor, AME, SE)

# 3) gewünschte Reihenfolge festlegen -----------------------------------
desired_order <- c(
  # Motivation
  "Political interest (z)",
  "Trust in parliament (z)",
  "Trust in federal council (z)",
  # Education
  "Compulsory education (vs. high)",
  "Secondary education (vs. high)",
  "Unknown education level (vs. high)",
  # Income
  "Low household income (vs. high)",
  "Middle household income (vs. high)",
  "Unknown household income (vs. high)",
  # Employment
  "Part-time employment (vs. full-time)",
  "Other employment status (vs. full-time)",
  "Unknown employment status (vs. full-time)",
  # Media & Engagement
  "Media use index (z)",
  "Political discussion (z)",
  "Party membership (yes vs. no)"
)

# 4) Predictor als Faktor und sortieren ----------------------------------
tbl_long <- tbl_long %>%
  mutate(Predictor = factor(Predictor, levels = desired_order)) %>%
  arrange(Model, Outcome, Predictor)

# 5) Tabelle ausgeben ----------------------------------------------------
kbl(tbl_long,
    caption   = "Average Marginal Effects , Standard Error by Model and Outcome in percentage points",
    booktabs  = TRUE,
    longtable = TRUE,
    align     = c("l","l","l","c","c"),
    col.names = c("Model","Outcome","Predictor","AME (pp)","SE (pp)")
) %>%
  kable_styling(
    latex_options = c("repeat_header","hold_position"),
    font_size     = 8,
    full_width    = FALSE
  ) %>%
  group_rows("Model 1 (Motivation)", 1,
             sum(tbl_long$Model == "Model 1")) %>%
  group_rows("Model 2 (+ Resources)",
             sum(tbl_long$Model == "Model 1") + 1,
             sum(tbl_long$Model %in% c("Model 1","Model 2"))) %>%
  group_rows("Model 3 (+ Media & Engagement)",
             sum(tbl_long$Model %in% c("Model 1","Model 2")) + 1,
             nrow(tbl_long)) %>%
  column_spec(3, width = "4cm") %>%
  footnote(
    general = c(
      "AME = average change in predicted probability when the predictor increases by one unit (in percentage points).",
      "SE = standard error; 95% CI (Confidence Interval) = AME ± 1.96 × SE."
    ),
    threeparttable    = TRUE,
    footnote_as_chunk = TRUE
  )


```

## Variable List

```{r}
library(dplyr)
library(knitr)
library(kableExtra)

# 1) Build metadata tibble ---------------------------------------------
vars <- tribble(
  ~Model,      ~Variable,          ~Label,                                                      ~Type,          ~Notes,
  "All",       "vote_group",        "Vote grouping (ref = “voter”)",                              "Factor (4)",   "voter / apathetic / protest / satisfied",
  
  "Model 1",   "edu_group",         "Education level",                                           "Factor (3)",   "low / mid / high",
  "Model 1",   "income_cat",        "Household income",                                          "Factor (3)",   "low / mid / high",
  "Model 1",   "job_cat",           "Employment status",                                         "Factor (3)",   "fulltime / parttime / other",

  "Model 2",   "media_index_z",     "Media-use index",                                           "Numeric (z)",  "Mean of 5 items; z-standardized",
  
  "Model 3",   "membership_dummy",  "Party-membership dummy",                                    "Integer (0/1)", "1 = member (codes 10–350); 0 = otherwise",
  "Model 3",   "discussion_z",      "Political-discussion frequency",                            "Numeric (z)",  "Wave 2 measure; z-standardized",
  "Model 3",   "w1_f14000",        "Closeness to a party",                                      "Integer (0/1)", "Codes 3–4 vs 1–2",

  "Auxiliary", "trust_parliament_z","Trust in parliament",                                       "Numeric (z)",  "10-point scale; z-standardized",
  "Auxiliary", "trust_council_z",   "Trust in federal council",                                  "Numeric (z)",  "10-point scale; z-standardized"
)

# 2) Render as grouped longtable ---------------------------------------
vars %>%
  arrange(factor(Model, levels = c("All","Model 1","Model 2","Model 3","Auxiliary"))) %>%
  group_by(Model) %>%
  do({
    df <- .
    if (first(df$Model) != "All") {
      df <- bind_rows(tibble(Model="",Variable="",Label="",Type="",Notes=""), df)
    }
    df
  }) %>%
  ungroup() %>%
  select(-Model) %>%
  kable(
    format       = "latex",
    booktabs     = TRUE,
    longtable    = TRUE,
    caption      = "Variable list for multinomial logit models",
    align        = c("l","l","l","l"),
    col.names    = c("Variable","Label","Type","Notes")
  ) %>%
  kable_styling(
    latex_options   = c("repeat_header","hold_position"),
    font_size       = 9,
    full_width      = FALSE
  ) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "4cm") %>%
  column_spec(3, width = "2.5cm") %>%
  column_spec(4, width = "5cm")

```

```{r}
library(dplyr)
library(knitr)
library(kableExtra)

# 1) Define metadata for classification variables --------------------------------
class_vars <- tribble(
  ~Group,              ~Variable,          ~Label,                                                            ~Type,           ~Coding,
  "Protest score",     "w3_f11304",        "Not convinced by any party/candidate",                            "Logical",       "1 = yes",
  "Protest score",     "w3_f12800a",       "Trust in federal council",                                        "Numeric",       "1–10; counted if ≤4 or =0",
  "Protest score",     "w3_f12800b",       "Trust in federal assembly",                                       "Numeric",       "1–10; counted if ≤4 or =0",
  "Protest score",     "w3_f11309",        "Elections do not change anything",                                "Logical",       "1 = yes",
  "Protest score",     "w3_f12800c",       "Trust in national parties",                                       "Numeric",       "1–10; counted if ≤4 or =0",
  "Protest score",     "w1_f13700",        "Satisfaction with democracy",                                     "Numeric",       "1–10; counted if ≤2 or =0",
  
  "Apathetic score",   "w3_f11303",        "Elections too complicated",                                       "Logical",       "1 = yes",
  "Apathetic score",   "w3_f11302",        "Not interested in politics",                                      "Logical",       "1 = yes",
  "Apathetic score",   "w3_f11306",        "Did not know candidates",                                         "Logical",       "1 = yes",
  "Apathetic score",   "w2_f10160",        "Paid little attention to campaign",                               "Logical",       "1 = yes",
  "Apathetic score",   "w3_f10100",        "Rather not interested in politics",                               "Numeric",       "1–4; counted if ≤2",
  
  "Satisfied score",   "w3_f12610a",       "Signed petition/initiative/referendum (past 5 yrs)",             "Logical",       "1 = yes",
  "Satisfied score",   "w3_f11302",        "Interested in politics",                                          "Logical",       "2 = yes",
  "Satisfied score",   "w1_f13700",        "Satisfaction with democracy",                                     "Numeric",       "1–10; counted if ≥3",
  "Satisfied score",   "w3_n13_030g2",     "Participated in party activities",                                "Logical",       "1 = yes",
  "Satisfied score",   "w3_n13_030e1",     "Member of environmental organization",                            "Logical",       "1 = yes",
  "Satisfied score",   "w3_n13_030d1",     "Member of human rights/peace organization",                       "Logical",       "1 = yes",
  "Satisfied score",   "w3_n13_030c1",     "Member of consumer organization",                                 "Logical",       "1 = yes",
  "Satisfied score",   "w3_n13_030b1",     "Member of economic/employers association",                        "Logical",       "1 = yes",
  "Satisfied score",   "w3_n13_030a1",     "Member of union/employee organization",                           "Logical",       "1 = yes",
  "Satisfied score",   "w3_f10100",        "Rather interested in politics",                                   "Numeric",       "1–4; counted if ≥3",
  "Satisfied score",   "w3_f12800c",       "Trust in national parties",                                       "Numeric",       "1–10; counted if ≥7 or =10",
  
  "Classification",    "score_protest",    "Sum of protest indicators",                                       "Integer",       "Range 0–6",
  "Classification",    "score_apathetic",  "Sum of apathetic indicators",                                     "Integer",       "Range 0–5",
  "Classification",    "score_satisfied",  "Sum of satisfied indicators",                                     "Integer",       "Range 0–10",
  "Classification",    "vote_group",       "Final group assignment",                                          "Factor",        "Based on max score; voters added separately"
)

# 2) Render grouped longtable --------------------------------------------------
class_vars %>%
  group_by(Group) %>%
  do({
    df <- .
    if (first(df$Group) != "Protest score") {
      df <- bind_rows(tibble(Group = "", Variable = "", Label = "", Type = "", Coding = ""), df)
    }
    df
  }) %>%
  ungroup() %>%
  select(-Group) %>%
  kable(
    format       = "latex",
    booktabs     = TRUE,
    longtable    = TRUE,
    caption      = "Variables used for non-voter group classification",
    align        = c("l","l","l","l","l"),
    col.names    = c("Variable","Label","Type","Coding")
  ) %>%
  kable_styling(
    latex_options = c("repeat_header","hold_position"),
    font_size     = 9,
    full_width    = FALSE
  ) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "5cm") %>%
  column_spec(3, width = "2.5cm") %>%
  column_spec(4, width = "4cm")

```

\newpage

## Declaration of AI Usage


Portions of this document were refined and enhanced using AI-assisted language tools to improve clarity, style, and overall readability. AI was also used as a coding help and coding inspiration. The substantive content and analysis remain the original work of the author.
\newpage

## R-Code
```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: FALSE
```
