---
title: "Non-voters in Switzerland"
subtitle: "A Data Analysis non voting"
author: 
  - "Felix Sigrist, felix.sigrist@stud.unilu.ch.com, 19-641-968"
  - "Supervisor: Prof. Dr. Benjamin Schlegel"
  - "Number of words: 2200"
lang: en
date: today
date-format: "D. MMMM YYYY"
format: 
  pdf:
    documentclass: scrreprt
    number-sections: true
    toc: true
    toc-depth: 2
    lof: false
    lot: false
    include-in-header:
      - text: |
          \usepackage{authblk}
          \usepackage{titling}
          \RedeclareSectionCommand[beforeskip=5pt plus 1pt minus 1pt]{section}
          \RedeclareSectionCommand[beforeskip=10pt plus 2pt minus 2pt]{chapter}
fontsize: 12pt
linestretch: 1.5
geometry: left=2cm,right=2cm,top=2.5cm,bottom=2.5cm
bibliography: nonvoting.bib
citeproc: true
execute:
  echo: false
  warning: false
knitr:
  opts_chunk:
    message: false
---

# Abstract {.unnumbered}

This paper examines how different types of non-voters in Switzerland can be classified and to which degree this classification models can predict whether someone votes or does not vote.

# Introduction

Although elections are a cornerstone of democratic legitimacy, more than half of the Swiss electorate regularly abstains from voting in federal elections. With turnout consistently below 50 percent since the 1970s [@figure1], abstention has become a structural feature of Swiss democracy [@freitag_labor_2005]. This raises important questions about democratic representation and political inclusion.

Despite its relevance, the phenomenon of non-voting has long been underexplored in Swiss political science. As Bühlmann, Freitag and Vatter [@vatter_pdf_2003] noted two decades ago, Swiss electoral research has shown “astonishing restraint” in addressing abstention empirically — even though Switzerland, together with the United States, consistently ranks at the bottom of voter turnout among established democracies [@ostrander_using_2021]. Figure 1 illustrates this trend, showing that turnout in federal elections has remained below 50 percent for over five decades.

Recent research has increasingly challenged the idea that abstention equals apathy. Typological approaches show that non-voters differ substantially in motivation, political attitudes and structural resources [@fatke_wollen_2015]. Studies on selective participation in Switzerland suggest that while some abstain situationally, others remain persistently disengaged [@dermont_taking_2016].

This study focuses on these **persistent non-voters** — individuals who abstained in both the 2019 and 2023 federal elections. It builds on the typology developed by Bühlmann et al. [@vatter_pdf_2003], condensing six original types into three analytically useful categories: **apathetic**, **protest-oriented**, and **satisfied or alternative non-voters**.

Using data from the SELECTS panel (2019–2023) and drawing on the Civic Voluntarism Model [@ostrander_using_2021], this study applies a multinomial logistic regression to explore how political attitudes, social integration and socioeconomic status explain persistent abstention — and how non-voters differ from the voting population and from each other.

```{r}
#| label: voter-turnout
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Figure 1: Voter turnout in federal elections has remained below 50% since the early 1970s — illustrating the long-term nature of electoral abstention in Switzerland."

library(ggplot2)

# Read in data
Participation <- read.csv("Daten/NRW-Zeitreihe.csv", sep = ";")

# Filter for national-level data
Participation_national <- subset(Participation, geoLevelName == "Schweiz")

# Convert turnout to numeric (handle commas)
Participation_national$wahlbeteiligung <- as.numeric(gsub(",", ".", Participation_national$wahlbeteiligung))

# Plot
ggplot(Participation_national, aes(x = wahl_jahr, y = wahlbeteiligung)) +
  geom_line(color = "#0072B2", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Voter Turnout in Swiss National Council Elections",
    x = "Election Year",
    y = "Turnout (%)"
  ) +
  theme_minimal(base_size = 13) +
  scale_x_continuous(breaks = seq(min(Participation_national$wahl_jahr), max(Participation_national$wahl_jahr), by = 4)) +
  ylim(30, 60)
 

```

# Literature Review

Political non-voting is a central theme in research on Swiss democracy. Several studies have aimed to typologize the heterogeneous group of non-voters and explain their behavior using theoretical models and empirical typologies.

## Classical Approaches to Political Participation

Earlier research on voter turnout was largely grounded in the **socio-structural** approach (e.g., Lazarsfeld), the **social-psychological** model (e.g., Campbell et al.), and **rational choice** theory (e.g., Downs). These were later extended through the **Standard Model of Political Participation** (Barnes & Kaase, 1979), which emphasized the role of socioeconomic resources and cognitive mobilization [@vatter_pdf_2003].

A conceptual advancement came with the introduction of the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which identifies three central factors influencing political participation: **resources**, **motivation**, and **recruitment networks**. The CVM has proven particularly useful in explaining unequal patterns of participation and in identifying different types of abstainers.

## Milbrath’s Hierarchical Model of Participation

Milbrath’s framework [@ruedin_testing_2007] conceptualizes political participation as a **cumulative hierarchy**—ranging from passive forms of involvement (e.g., voting, discussing politics) to more active and elite-level activities (e.g., running for office). This model combines individual characteristics with exposure to political stimuli and has been validated through simulation studies that demonstrate the dynamic interplay of individual and institutional factors.

## A Swiss Typology: Bühlmann et al. (2003)

[@vatter_pdf_2003] developed one of the first typologies of Swiss non-voters using cluster analysis on survey data from the 1999 federal elections. Their typology includes six distinct types:

-   Politically Disinterested\
-   Politically Disenchanted\
-   Socially Isolated\
-   Politically Incompetent\
-   Protest Voters\
-   Alternative Participants

These types differ in terms of political interest, efficacy, trust in institutions, and socioeconomic resources.

## Extension and Replication: Fatke & Freitag (2015)

[@fatke_wollen_2015] built upon this typology using SELECTS 2011 data. Their cluster analysis refined the groupings and reaffirmed the heterogeneity of non-voters. Types such as *satisfied but disengaged*, *incompetent*, and *abstaining but participatory* highlight that abstention does not always stem from apathy or alienation, but also from rational disengagement or alternative political expression.

## Democratic Implications and Empirical Consensus

Across all studies, a key insight prevails: **non-voters are not a homogeneous group**. From a democratic theory perspective, the most problematic are the **socially isolated** and **incompetent**, who lack both the resources and social networks for political engagement [@fatke_nichtwahlenden_2015].

At the same time, not all abstention is a cause for concern. **Alternative participants** or the **satisfied disengaged**, for example, abstain not due to rejection or apathy, but because they either prefer other forms of participation (e.g., initiatives, petitions) or are fundamentally satisfied with the system.

This differentiated understanding provides the conceptual bridge to the following theory section, which formalizes these empirical distinctions into a testable framework.

------------------------------------------------------------------------

# Theory

## Electoral Abstention as a Multi-Dimensional Phenomenon

Political abstention is increasingly seen as a multidimensional phenomenon. Contemporary research emphasizes that non-voters differ systematically in terms of political motivation, cognitive engagement, social integration, and structural access. Typological approaches have shown that abstention reflects distinct profiles of political disengagement, rather than a binary act of participation versus non-participation [@vatter_pdf_2003; @fatke_wollen_2015].

## The Civic Voluntarism Model as Theoretical Foundation

This study applies the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which explains political engagement through three core components:

1.  **Resources** – time, money, civic skills, and knowledge\
2.  **Motivation** – interest in politics, efficacy, ideological conviction\
3.  **Recruitment Networks** – social integration, group membership, mobilization

According to CVM, individuals are more likely to participate politically when they are motivated, able, and socially embedded. Conversely, the absence of any one of these factors can lead to abstention. Crucially, access to these dimensions is unequally distributed—meaning that political inequality is reproduced through structural exclusion [@ostrander_using_2021].

The CVM therefore provides a robust foundation for explaining **persistent abstention**, as it integrates individual capacities, perceptions, and social context into one explanatory model.

## Typologies of Non-Voters in Switzerland

Building on the CVM, the typologies developed by Bühlmann et al. [@vatter_pdf_2003] and extended by Fatke & Freitag [@fatke_wollen_2015] offer a detailed framework for classifying non-voters. For the purpose of this study, these types are aggregated into three theoretically meaningful categories:

| Type | Main Mechanism | Theoretical Foundation |
|----------------------|-------------------------|-------------------------|
| **Apathetic** | Lack of motivation/resources | Civic Voluntarism Model |
| **Protest-oriented** | Political alienation | System support/trust theory |
| **Satisfied/Alternative** | Rational disengagement | Expressive/instrumental logic |

This typology provides the conceptual basis for analyzing **persistent non-voters**, i.e., those who abstained in both 2019 and 2023 federal elections in Switzerland.

## Persistent Non-Voting as an Analytical Focus

Persistent abstention is of particular interest, as it suggests more entrenched and stable forms of disengagement. Unlike situational abstention, which may result from temporary constraints, persistent non-voters are likely to exhibit deeper structural or attitudinal barriers. Typologizing this group allows for a theory-driven understanding of *why* certain individuals consistently opt out of electoral politics.

## Theoretical Expectations and Hypotheses

Based on the typology and the Civic Voluntarism Model, the following hypotheses are derived:

-   **H1 (Political Interest & Knowledge):** Apathetic non-voters report significantly lower political interest and political knowledge than other groups.

-   **H2 (System Satisfaction & Trust):** Protest-oriented non-voters express lower trust in political institutions and lower satisfaction with democracy than other groups.

-   **H3 (Socioeconomic Status):** Apathetic non-voters have lower levels of education and income, while satisfied/alternative non-voters resemble voters in their socioeconomic status.

-   **H4 (Social Integration):** Apathetic non-voters show weaker social integration (e.g., fewer memberships or interpersonal networks) than other groups.

-   **H5 (Party Identification & Ideological Self-Placement):** Protest-oriented non-voters are ideologically positioned but report weak or no party identification.

These hypotheses will be empirically tested using a multinomial logistic regression model, treating **voters** as the reference category. The predictors reflect the theoretical dimensions discussed above.

# Empirical Analysis

## Data and Descriptive Statistics

The analysis is based on data from the Swiss Election Study (SELECTS) panel survey, covering the 2019 and 2023 federal elections. The dataset includes respondents who participated in all relevant waves and provides detailed information on political attitudes, behavior, and sociodemographic background.

To identify persistent abstainers, we selected respondents who reported not voting in both the 2019 and 2023 National Council elections. Voting behavior in 2019 is captured by the variable `w1_f10200rec` (re-coded: 1 = voted, 0 = did not vote), and in 2023 by `w3_f11100`. Respondents who reported voting in both elections were classified as **voters** and serve as the reference category in the analysis.

Persistent non-voters were further classified into three theoretical groups based on their stated reasons for abstention in 2023. The variable block `w3_f11301` to `w3_f11310` captures multiple reasons for non-voting. Following the typology developed by Bühlmann et al. (2003) and adapted by Fatke & Freitag (2015), we grouped persistent abstainers into the following types:

- **Apathetic**: cited reasons such as lack of interest, insufficient information, or complexity of the voting process.
- **Protest-oriented**: cited dissatisfaction with parties, politics, or the political system.
- **Satisfied/Alternative**: cited contentment with the system or a belief that their participation was not necessary.

Respondents who did not fall clearly into one of these categories were excluded from the analysis. The resulting dependent variable `vote_group` consists of four categories: `voter`, `apathetic`, `protest`, and `satisfied`.

Table 1 below summarizes the group sizes:

```{r}
#| label: table-votegroups
#| echo: false
#| results: asis

library(dplyr)
library(knitr)

df_model %>%
  count(vote_group, name = "N") %>%
  arrange(desc(N)) %>%
  kable(caption = "Table 1: Distribution of vote groups in the analytical sample",
        col.names = c("Group", "N"))


```

```{r}
# Load required packages
library(tidyverse)
library(haven)
library(janitor)
library(naniar)

# 1. Load and clean the SPSS dataset
data_path <- "Daten/2626_Selects2023_Panel_Data_v2.0.sav"

df <- read_sav(data_path) %>%
  zap_labels() %>%           # remove SPSS labels
  clean_names()              # snake_case variable names

# 2. Identify persistent non-voters (abstained in both 2019 and 2023)
df_nv <- df %>%
  filter(w1_f10200rec == 0,            # did not vote in 2019
         w3_f11100 %in% c(1, 2, 3))    # did not vote in 2023 (various forms)

# 3. Classify non-voters into types based on reasons for abstention in 2023
df_nv <- df_nv %>%
  mutate(
    vote_group = case_when(
      # Apathetic: no interest, too complicated, no information
      w3_f11301 == 1 | w3_f11302 == 1 | w3_f11303 == 1 ~ "apathetic",

      # Protest-oriented: system or party dissatisfaction
      w3_f11306 == 1 | w3_f11307 == 1 ~ "protest",

      # Satisfied/Alternative: no need to vote, generally satisfied
      w3_f11309 == 1 | w3_f11310 == 1 ~ "satisfied",

      TRUE ~ NA_character_
    )
  )

# 4. Add the group of consistent voters (voted in both elections)
df <- df %>%
  mutate(
    vote_group = case_when(
      w1_f10200rec == 1 & w3_f11100 == 1 ~ "voter",
      TRUE ~ NA_character_
    )
  )

# 5. Merge back the classified non-voters into the main dataset
df$vote_group[df$id %in% df_nv$id] <- df_nv$vote_group

# 6. Optional: Filter final dataset to only include valid vote_group values
df_model <- df %>%
  filter(!is.na(vote_group))

# 7. Check group counts
table(df_model$vote_group, useNA = "always")


```



```{r}
library(nnet)
library(broom.mixed)
library(jtools)
library(haven)

df<-read_sav("Daten/2626_Selects2023_Panel_data_v2.0.sav")

# Modell 1: Politisches Interesse & Systemvertrauen
model1 <- multinom(vote_group ~ pol_interest + trust_parliament + trust_government, data = df)

# Übersicht
summary(model1)

# P-Werte berechnen (standardmäßig nicht enthalten)
z <- summary(model1)$coefficients / summary(model1)$standard.errors
p <- 2 * (1 - pnorm(abs(z)))

# Ergebnis-Tabelle (log-odds)
round(p, 4)

# Alternativ: nicer Output
library(jtools)
summ(model1, exp = TRUE)  # exp=TRUE zeigt Odds Ratios statt Logits
```

## Methods

We estimate an OLS model and a logistic regression with interaction terms. The dependent variable is voting for the incumbent, and key predictors include economic evaluations and political information.

## Results

OLS results are shown in @tbl-reg.

Logistic regression results explore interaction effects, visualized in @fig-interaction.

# Discussion

Lorem

# Conclusion

Lorem
