---
title: "Beyond Apathy: Structural and Motivational Drivers of Electoral Abstention"
subtitle: "A Data Analysis non voting"
author: 
  - "Felix Sigrist, felix.sigrist@stud.unilu.ch.com, 19-641-968"
  - "Supervisor: Prof. Dr. Benjamin Schlegel"
  - "Number of words: 4284"
lang: en
date: today
date-format: "D. MMMM YYYY"
format: 
  pdf:
    documentclass: scrreprt
    number-sections: true
    toc: true
    toc-depth: 2
    lof: false
    lot: false
    include-in-header:
      - text: |
          \usepackage{authblk}
          \usepackage{titling}
          \RedeclareSectionCommand[beforeskip=5pt plus 1pt minus 1pt]{section}
          \RedeclareSectionCommand[beforeskip=10pt plus 2pt minus 2pt]{chapter}
fontsize: 12pt
linestretch: 1.5
geometry: left=2cm,right=2cm,top=2.5cm,bottom=2.5cm
bibliography: nonvoting.bib
citeproc: true
execute:
  echo: false
  warning: false
knitr:
  opts_chunk:
    message: false
---

# Abstract {.unnumbered}

### Abstract

This research note investigates persistent non-voters in Switzerland using panel data from the SELECTS 2019–2023 study. Building on existing typologies, the analysis distinguishes between apathetic, protest-oriented, and satisfied non-voters. A multinomial logistic regression model tests the influence of political interest, institutional trust, socioeconomic status, and social integration. Results show that persistent abstention is primarily driven by motivational and structural factors. Apathetic non-voters are less politically engaged and socioeconomically disadvantaged, while satisfied abstainers resemble voters but opt out for normative reasons. Protest abstention plays a minor role. The findings highlight the need for differentiated engagement strategies tailored to the distinct profiles of non-voters.


# Introduction

Although elections are a cornerstone of democratic legitimacy, more than half of the Swiss electorate regularly abstains from voting in federal elections. With turnout consistently below 50 percent since the 1970s (@fig-voter-turnout), abstention has become a structural feature of Swiss democracy [@freitag_labor_2005]. This raises important questions about democratic representation and political inclusion.

Despite its relevance, the phenomenon of non-voting has long been underexplored in Swiss political science. As Bühlmann, Freitag and Vatter [@vatter_pdf_2003] noted two decades ago, Swiss electoral research has shown “astonishing restraint” in addressing abstention empirically — even though Switzerland, together with the United States, consistently ranks at the bottom of voter turnout among established democracies [@ostrander_using_2021]. @fig-voter-turnout illustrates this trend, showing that turnout in federal elections has remained below the 50 percent mark for over five decades.

Recent research has increasingly challenged the idea that abstention equals apathy. Typological approaches show that non-voters differ substantially in motivation, political attitudes and structural resources [@fatke_wollen_2015]. Studies on selective participation in Switzerland suggest that while some abstain situationally, others remain constantly disengaged [@dermont_taking_2016].

This study focuses on these **persistent non-voters** — individuals who abstained in both the 2019 and 2023 federal elections. It builds on the typology developed by Bühlmann et al. [@vatter_pdf_2003], condensing six original types into three analytically useful categories: **apathetic**, **protest-oriented**, and **satisfied or alternative non-voters**.

Using data from the SELECTS panel (2019–2023) and drawing on the Civic Voluntarism Model [@ostrander_using_2021], this study applies a multinomial logistic regression to explore how political attitudes, social integration and socioeconomic status explain persistent abstention — and how non-voters differ from the voting population and from each other.

```{r}
#| label: "fig-voter-turnout"
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Voter turnout in federal elections has remained below 50% since the early 1970s — illustrating the long-term nature of electoral abstention in Switzerland. Data source: [@noauthor_wahlbeteiligung_nodate]."

library(ggplot2)

# Read in data
Participation <- read.csv("Daten/NRW-Zeitreihe.csv", sep = ";")

# Filter for national-level data
Participation_national <- subset(Participation, geoLevelName == "Schweiz")

# Convert turnout to numeric (handle commas)
Participation_national$wahlbeteiligung <- as.numeric(gsub(",", ".", Participation_national$wahlbeteiligung))

# Plot
ggplot(Participation_national, aes(x = wahl_jahr, y = wahlbeteiligung)) +
  geom_line(color = "#0072B2", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Voter Turnout in Swiss National Council Elections",
    x = "Election Year",
    y = "Turnout (%)"
  ) +
  theme_minimal(base_size = 13) +
  scale_x_continuous(breaks = seq(min(Participation_national$wahl_jahr), max(Participation_national$wahl_jahr), by = 4)) +
  ylim(30, 60)
 

```

# Literature Review

Political non-voting is a central theme in research on Swiss democracy. Several studies have aimed to typologize the heterogeneous group of non-voters and explain their behavior using theoretical models and empirical typologies.

## Classical Approaches to Political Participation

Earlier research on voter turnout was largely grounded in the **socio-structural** approach (e.g., Lazarsfeld), the **social-psychological** model (e.g., Campbell et al.), and **rational choice** theory (e.g., Downs). These were later extended through the **Standard Model of Political Participation** (Barnes & Kaase, 1979), which emphasized the role of socioeconomic resources and cognitive mobilization [@vatter_pdf_2003].

A conceptual advancement came with the introduction of the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which identifies three central factors influencing political participation: **resources**, **motivation**, and **recruitment networks**. The CVM has proven particularly useful in explaining unequal patterns of participation and in identifying different types of abstainers (ebd.).

## Milbrath’s Hierarchical Model of Participation

Milbrath’s framework [@ruedin_testing_2007] conceptualizes political participation as a **cumulative hierarchy**—ranging from passive forms of involvement (e.g., voting, discussing politics) to more active and elite-level activities (e.g., running for office). This model combines individual characteristics with exposure to political stimuli and has been validated through simulation studies that demonstrate the dynamic interplay of individual and institutional factors (ebd.).

## A Swiss Typology: Bühlmann et al. (2003)

[@vatter_pdf_2003] developed one of the first typologies of Swiss non-voters using cluster analysis on survey data from the 1999 federal elections. Their typology includes six distinct types:

-   Politically Disinterested\
-   Politically Disenchanted\
-   Socially Isolated\
-   Politically Incompetent\
-   Protest Voters\
-   Alternative Participants

These types differ in terms of political interest, efficacy, trust in institutions, and socioeconomic resources.

## Extension and Replication: Fatke & Freitag (2015)

[@fatke_wollen_2015] built upon this typology using SELECTS 2011 data. Their cluster analysis refined the groupings and reaffirmed the heterogeneity of non-voters. Types such as *satisfied but disengaged*, *incompetent*, and *abstaining but participatory* highlight that abstention does not always stem from apathy or alienation, but also from rational disengagement or alternative political expression.

## Democratic Implications and Empirical Consensus

Across all studies, a key insight prevails: **non-voters are not a homogeneous group**. From a democratic theory perspective, the most problematic are the **socially isolated** and **incompetent**, who lack both the resources and social networks for political engagement [@fatke_nichtwahlenden_2015].

At the same time, not all abstention is a cause for concern. **Alternative participants** or the **satisfied disengaged**, for example, abstain not due to rejection or apathy, but because they either prefer other forms of participation (e.g., initiatives, petitions) or are fundamentally satisfied with the system.

This differentiated understanding provides the conceptual bridge to the following theory section, which formalizes these empirical distinctions into a testable framework.

------------------------------------------------------------------------

# Theory

## Theoretical framework

### The Civic Voluntarism Model as Theoretical Foundation

This study applies the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which explains political engagement through three core components:

1.  **Resources** – time, money, civic skills, and knowledge\
2.  **Motivation** – interest in politics, efficacy, ideological conviction\
3.  **Recruitment Networks** – social integration, group membership, mobilization

According to CVM, individuals are more likely to participate politically when they are motivated, able, and socially embedded. Conversely, the absence of any one of these factors can lead to abstention. Crucially, access to these dimensions is unequally distributed—meaning that political inequality is reproduced through structural exclusion [@ostrander_using_2021].

The CVM therefore provides a robust foundation for explaining **persistent abstention**, as it integrates individual capacities, perceptions, and social context into one explanatory model.

### Electoral Abstention as a Multi-Dimensional Phenomenon

Political abstention is increasingly seen as a multidimensional phenomenon. Contemporary research emphasizes that non-voters differ systematically in terms of political motivation, cognitive engagement, social integration, and structural access. Typological approaches have shown that abstention reflects distinct profiles of political disengagement, rather than a binary act of participation versus non-participation [@vatter_pdf_2003; @fatke_wollen_2015].

### Persistent Non-Voting as an Analytical Focus

Persistent abstention is of particular interest, as it suggests more entrenched and stable forms of disengagement. Unlike situational abstention, which may result from temporary constraints, persistent non-voters are likely to exhibit deeper structural or attitudinal barriers. Typologizing this group allows for a theory-driven understanding of *why* certain individuals consistently opt out of electoral politics.

## Typologies of Non-Voters in Switzerland

While the Civic Voluntarism Model explains many forms of non-participation through resource or motivation deficits, protest-oriented abstention requires an additional conceptual lens. Drawing on theories of political support and system trust \[Easton 1975; Norris 2011\], political scientists argue that abstention may also reflect alienation from democratic institutions. In this view, protest abstention is not a symptom of apathy but rather a passive expression of dissatisfaction — a silent refusal to endorse a political system perceived as unresponsive, unfair, or corrupt.

This distinction is particularly relevant in the Swiss context, where institutional trust is generally high, but perceived political inefficacy may still motivate strategic disengagement. Incorporating this dimension allows for a clearer differentiation between those who are disengaged and those who abstain deliberately out of protest.

Building on the CVM and empiric findings, this section presents the typology of persistens non-voters which serves as the dependent variable in this paper. The typologies developed by Bühlmann et al. [@vatter_pdf_2003] and extended by Fatke & Freitag [@fatke_wollen_2015] offer a detailed framework for classifying non-voters. For the purpose of this study, these types are aggregated into three theoretically meaningful categories:

| Type | Main Mechanism | Theoretical Foundation |
|------------------------|------------------------|------------------------|
| **Apathetic** | Lack of motivation/resources | Civic Voluntarism Model |
| **Protest-oriented** | Political alienation | System support/trust theory |
| **Satisfied/Alternative** | Rational disengagement | Expressive/instrumental logic |

This typology provides the conceptual basis for analyzing **persistent non-voters**, i.e., those who abstained in both 2019 and 2023 federal elections in Switzerland.

## Theoretical Expectations and Hypotheses

In order to explain the differences betrween the different types of abstention, testabale hypotheses are derived from the Civic Voluntarism Model and the typology shown above.

-   **H1 (Political Interest & Knowledge):** Apathetic non-voters report significantly lower political interest and political knowledge than other groups.

-   **H2 (System Satisfaction & Trust):** Protest-oriented non-voters express lower trust in political institutions and lower satisfaction with democracy than other groups.

-   **H3 (Socioeconomic Status):** Apathetic non-voters have lower levels of education and income, while satisfied/alternative non-voters resemble voters in their socioeconomic status.

-   **H4 (Social Integration):** Apathetic non-voters show weaker social integration (e.g., fewer memberships or interpersonal networks) than other groups.

-   **H5 (Party Identification & Ideological Self-Placement):** Protest-oriented non-voters are ideologically positioned but report weak or no party identification.

These hypotheses will be empirically tested using a multinomial logistic regression model, treating **voters** as the reference category. The predictors reflect the theoretical dimensions discussed above.

# Empirical Analysis

## Data and Descriptive Statistics

The analysis is based on data from the Swiss Election Study (SELECTS) panel survey, covering the 2019 and 2023 federal elections. The dataset includes respondents who participated in all relevant waves and provides detailed information on political attitudes, behavior, and sociodemographic background.

To identify persistent abstainers, respondents who reported not voting in both the 2019 and 2023 National Council elections were selected. Voting behavior in 2019 is captured by the variable `w1_f10200rec` (re-coded: 1 = voted, 0 = did not vote), and in 2023 by `w3_f11100`. Respondents who reported voting in both elections were classified as **voters** and serve as the reference category in the analysis.

Persistent non-voters were further classified into three theoretical groups based on their stated reasons for abstention in 2023. The variable block `w3_f11301` to `w3_f11310` captures multiple reasons for non-voting. Following the typology developed by Bühlmann et al. (2003) and adapted by Fatke & Freitag (2015), persistent abstainers were grouped into the following types:

-   **Apathetic**: cited reasons such as lack of interest, insufficient information, or complexity of the voting process.
-   **Protest-oriented**: cited dissatisfaction with parties, politics, or the political system.
-   **Satisfied/Alternative**: cited contentment with the system or a belief that their participation was not necessary.

Respondents who did not fall clearly into one of these categories were excluded from the analysis. The resulting dependent variable `vote_group` consists of four categories: `voter`, `apathetic`, `protest`, and `satisfied`.

Table 4.1 below summarizes the group sizes:

```{r}
#| echo: false
#| results: false
library(tidyverse)
library(haven)
library(janitor)
library(naniar)

# Load and clean dataset
df <- read_sav("Daten/2626_Selects2023_Panel_Data_v2.0.sav") %>%
  zap_labels() %>%           
  clean_names()              

# Identify persistent non-voters (did not vote in both 2019 and 2023)
df_nv <- df %>%
  filter(w1_f10200rec == 0, w3_f11100 %in% c(1, 2, 3))

# Classify non-voters into types based on 2023 reasons
df_nv <- df_nv %>%
  mutate(
    vote_group = case_when(
      w3_f11301 == 1 | w3_f11302 == 1 | w3_f11303 == 1 ~ "apathetic",
      w3_f11306 == 1 | w3_f11307 == 1 ~ "protest",
      w3_f11309 == 1 | w3_f11310 == 1 ~ "satisfied",
      TRUE ~ NA_character_
    )
  )

# Add consistent voters
df <- df %>%
  mutate(
    vote_group = case_when(
      w1_f10200rec == 1 & w3_f11100 == 4 ~ "voter",
      TRUE ~ NA_character_
    )
  )

# Merge voter types back into main dataset
df$vote_group[df$id %in% df_nv$id] <- df_nv$vote_group

# Create all additional variables BEFORE filtering
# Create all additional variables BEFORE filtering
df <- df %>%
  mutate(
    # Z-standardized predictors
    interest_z = scale(w3_f10100)[,1],
    trust_council_z = scale(w3_f12800a)[,1],    # Federal Council
    trust_parliament_z = scale(w3_f12800b)[,1], # Parliament
    discussion_z = scale(w2_f13310)[,1],        # from Wave 2

    # Media use index from 5 sources (Wave 3)
    media_index = rowMeans(select(., w3_f12610a, w3_f12610b, w3_f12610d, w3_f12610g, w3_f12610i), na.rm = TRUE),
    media_index_z = scale(media_index)[,1],

    # Dummy: party membership (1 = yes, 0 = no)
    membership_dummy = if_else(w3_f14081rec == 1, 1, 0),

    # Categorical predictors
    vote_group = relevel(factor(vote_group), ref = "voter"),
    education = w1_f21310rec,
    income = w1_f28910,
    employment = w1_f21400,

    # Grouped categories
    edu_group = case_when(
      education %in% c(1, 2, 3) ~ "low",
      education %in% c(4, 5, 6) ~ "mid",
      education %in% c(7, 8, 9, 10, 11, 12) ~ "high",
      TRUE ~ "unknown"
    ),
    income_cat = case_when(
      income %in% c(1, 2, 3, 4) ~ "low",
      income %in% c(5, 6, 7) ~ "mid",
      income %in% c(8, 9, 10, 11) ~ "high",
      TRUE ~ "unknown"
    ),
    job_cat = case_when(
      employment == 1 ~ "fulltime",
      employment == 2 ~ "parttime",
      employment %in% c(3, 4, 5, 6, 7, 8) ~ "other",
      TRUE ~ "unknown"
    )
  )


# Final filtered dataset for modelling
df_model <- df %>%
  filter(!is.na(vote_group))

# Show group distribution
table(df_model$vote_group, useNA = "always")


```

```{r}
#| label: table-votegroups
#| echo: false
#| results: asis
#| warning: false
#| message: false

library(dplyr)
library(knitr)

df_model %>%
  count(vote_group, name = "N") %>%
  arrange(desc(N)) %>%
  mutate(share = round(100 * N / sum(N), 1)) %>%
  kable(
    caption = "Distribution of vote groups in the analytical sample",
    col.names = c("Group", "N", "Share (%)")
  )

```

The distribution confirms that **apathetic non-voters** represent the largest group in the analytical sample, followed by **protest-oriented** and **satisfied** non-voters. Notably, about one third of the sample are consistent voters. This distribution reflects the empirical prevalence of non-voting types among persistent abstainers and underlines the heterogeneity of non-voter profiles. This also suggests that motivational and resource-related non-voting could be more widespread than ideologically motivated protest or strategic satisfaction.

## Methods

To test the proposed hypotheses, we estimate a multinomial logistic regression model. This modeling approach is appropriate for nominal dependent variables with more than two categories. The dependent variable in this analysis is `vote_group`, which consists of four categories: **voters** (reference category), **apathetic**, **protest-oriented**, and **satisfied/alternative** non-voters.

The general form of the multinomial logistic regression model is:

$$
\log\left( \frac{P(Y = k)}{P(Y = \text{Voter})} \right) = \beta_{0k} + \beta_{1k}X_1 + \beta_{2k}X_2 + \dots + \beta_{pk}X_p, \quad \text{for } k \in \{\text{Apathetic, Protest, Satisfied}\}
$$

Here, \(Y\) denotes the categorical outcome variable (`vote_group`), and the coefficients \(\beta_{ik}\) represent the effect of predictor \(X_i\) on the log-odds of being in category \(k\) compared to the voter reference category.

The analysis is conducted in three steps, each corresponding to a model with increasing complexity:

- **Model 1** includes political interest and political trust — key motivational components from the Civic Voluntarism Model.
- **Model 2** adds indicators of socioeconomic status: education level, income, and occupational status.
- **Model 3** further includes variables related to social integration and political communication: media use, organizational membership, and frequency of political discussions.

All models are estimated using **multinomial logistic regression** with robust standard errors. The interpretation of coefficients is based on **log-odds** relative to the reference category (**voters**). For ease of interpretation, average marginal effects (AMEs) are reported in the results section.

Prior to analysis, all continuous predictors are standardized (mean = 0, sd = 1). Categorical predictors are converted to factors. Missing values are handled using listwise deletion. The final analytical sample consists of 589 cases with a valid `vote_group` classification and complete data on the relevant predictors.


## Variable Operationalization

The dependent variable `vote_group` is a categorical variable with four values: **voter** (reference category), **apathetic**, **protest-oriented**, and **satisfied/alternative** non-voters. It is constructed based on self-reported turnout behavior in 2019 (`w1_f10200rec`) and 2023 (`w3_f11100`), as well as respondents’ stated reasons for abstaining in 2023 (`w3_f11301` to `w3_f11310`).

The independent variables are grouped according to the logic of the Civic Voluntarism Model and introduced step-by-step in three models:

### Model 1: Political Motivation

The first model captures motivational factors as conceptualized in the Civic Voluntarism Model. It includes attitudinal variables that reflect individuals’ political interest and trust in core democratic institutions:

- **Political Interest** (`w3_f10100`): Self-reported interest in politics on a scale from 0 (not at all interested) to 10 (very interested).
- **Trust in Parliament** (`w3_f12800a`): Level of trust in the National Parliament (0–10 scale).
- **Trust in Federal Council** (`w3_f12800b`): Level of trust in the Federal Council (0–10 scale).

All variables are treated as continuous and standardized prior to inclusion in the model.



### Model 2: Socioeconomic Resources

The second model introduces structural indicators of socioeconomic status, representing the “resources” dimension of the Civic Voluntarism Model. These variables are operationalized as follows:

- **Education Level** (`w3_f21310`): Highest level of completed education, recoded to an ordinal scale (from compulsory schooling to tertiary education).
- **Income** (`w3_f22000_rec`): Monthly household income, recoded into 11 ordered categories.
- **Employment Status** (`w3_f21000`): Employment situation (e.g., full-time, unemployed, student), recoded into a categorical factor.

All predictors are entered as ordinal or categorical variables. This model tests whether socioeconomic disadvantages are associated with persistent abstention.


### Model 3: Social Integration & Political Communication

In the third model, the analysis includes variables which are related to social integration, political communication and simplified indicators of socioeconomic status to ensure stable and interpretable estimates. Due to sparse cell counts and limited observations in some of the subcategories, a data-driven simplification of the categorial variables for education, income and employment was conducted as follows:

-   **Education Level** (`edu_group_simple`): This variable was originally differentiated into multiple groups, due to the low number of cases it was reduced to two categories:

    -   `teritary:`Respondents with higher education

    -   `non-teritary:` Respondents with medium, low, or unknown educational levels.

-   **Income** (`income_simple`): Due to sparsity in low and middle-income categories among non-voters, the original income variable was simplified into two categories:

    -   `high:`Respondents with high income

    -   `low-mid:` Respondents combining low, middle and unknown income levels.

-   **Employment** (`job_simple`): Employment categories were similarly condensed into two groups:

    -   `fulltime:`Respondents employed full-time.

    -   `low-mid:` Respondents employed part-time, unemployed, retired, or with an unknown employment status.

Additionally, the following variables on political communication and integration remain included in their original, detailed form:

-   **Media Use** (`media_index_z`): Aggregated and standardized frequency of political news consumption via various media (newspapers, TV, internet)

-   **Political Discussion Frequency** (`discussion_z`): Standardized measure of how frequently respondents discuss politics with friends or family (5-point scale).

## Results

```{r load-models-safe, echo=FALSE, message=FALSE, warning=FALSE}
library(nnet)

# Lade Modelle mit Fehlerprüfung
model1 <- try(readRDS("models/model1.rds"), silent = TRUE)
model2_clean <- try(readRDS("models/model2_clean.rds"), silent = TRUE)
model3_better <- try(readRDS("models/model3_better.rds"), silent = TRUE)

# Bei Fehler abbrechen
if (inherits(model1, "try-error") || !inherits(model1, "multinom")) stop("❌ 'model1' ist kein gültiges Modell.")
if (inherits(model2_clean, "try-error") || !inherits(model2_clean, "multinom")) stop("❌ 'model2_clean' ist kein gültiges Modell.")
if (inherits(model3_better, "try-error") || !inherits(model3_better, "multinom")) stop("❌ 'model3_better ist kein gültiges Modell.")

# Z- und P-Werte berechnen
z1 <- summary(model1)$coefficients / summary(model1)$standard.errors
p1 <- 2 * (1 - pnorm(abs(z1)))

z2_clean <- summary(model2_clean)$coefficients / summary(model2_clean)$standard.errors
p2_clean <- 2 * (1 - pnorm(abs(z2_clean)))

z3_better <- summary(model3_better)$coefficients / summary(model3_better)$standard.errors
p3_better <- 2 * (1 - pnorm(abs(z3_better)))

```

```{r stargazer-wide-fix, echo=FALSE, results='asis'}
library(stargazer)

cat("\\begin{table}[H]\n\\centering\n\\resizebox{\\textwidth}{!}{%\n")

# Stargazer ohne Float
stargazer(
  model1, model2_clean, model3_better,
  type = "latex",
  float = FALSE,
  style = "default",
  title = "",
  dep.var.labels = c("Reference category: Voter"),
  column.labels = c("Model 1", "Model 2 (grouped)", "Model 3 (full)"),
  covariate.labels = c(
    "Political Interest", 
    "Trust in Parliament", 
    "Trust in Federal Council",
    "Education: Secondary",
    "Education: Tertiary",
    "Income: Mid",
    "Income: High",
    "Employment: Part-time",
    "Employment: Unemployed",
    "Employment: Retired",
    "Media Use", 
    "Party Membership", 
    "Political Discussion"
  ),
  omit.stat = c("ll", "aic", "bic"),
  no.space = TRUE,
  star.cutoffs = c(0.05, 0.01, 0.001),
  notes = c("Robust standard errors in parentheses", "* p < 0.05; ** p < 0.01; *** p < 0.001")
)

cat("}\n\\caption{Multinomial logistic regression comparing non-voter types with voters as the reference category.}\n\\end{table}")
```

The results from the three multinomial logistic regression models are summarized in Table X (using `stargazer`). The models test a theory-driven typology of non-voters by examining whether motivational, structural, and integrative factors predict group membership relative to consistent voters.

**Model 1** (H1, H2) shows that political interest is a highly significant predictor across all non-voter types. This supports **Hypothesis 1**, which posited that apathetic non-voters are less interested in politics than voters. Trust in the national parliament shows a modest effect—primarily for the apathetic group—while trust in the federal council remains insignificant. These findings offer partial support for **Hypothesis 2**, suggesting that protest-oriented non-voters are not simply marked by low trust in government institutions.

**Model 2** (H3a, H3b) adds education, income, and employment as predictors and improves overall model fit (ΔAIC = -30). Apathetic non-voters are clearly linked to lower levels of education and income, confirming **Hypothesis 3a**. Meanwhile, satisfied non-voters appear socioeconomically closer to voters, consistent with **Hypothesis 3b**. The employment variable, however, shows no consistent pattern.

**Model 3** (H4, H5) integrates indicators of social integration and political communication: media use, political discussion and party membership. To adress previous estimation problems due to sparse data, a theoretically grounded recoding and grouped vategories for education, income and employment were set in place. Despite these corrections, none of the additional predictors in Model 3 reach statistical significance. As a result, **Hypothesis 4**, which expected lower media exposure among apathetic non-voters, is not supported. Similarly, **Hypothesis 5**, which assumed weaker social integration among apathetic and protest-oriented abstainers, receives no empirical support. 

Altough Model 3 was originally not intended to be a robustness check, its lack of significatn new findings shifts its interpretive value. Rather than offering further explanatory dimensions, it strenghtens the conclusions from Models 1 and 2: Persistent abstention in Switzerland is less shaped by media environments or network embeddedness and more by fundamental differences in political motivation and structural inequality. 

```{r}
#| label: table-hypotheses-clean
#| echo: false
#| message: false
#| warning: false
#| results: asis

library(dplyr)
library(knitr)

# Hypotheses table without symbols
hypotheses_clean <- tibble::tibble(
  Hypothesis = c("H1", "H2", "H3a", "H3b", "H4", "H5"),
  Description = c(
    "Apathetic non-voters show lower political interest than voters",
    "Protest non-voters show lower trust and satisfaction",
    "Apathetic non-voters have lower socioeconomic status",
    "Satisfied non-voters resemble voters socioeconomically",
    "Apathetic non-voters consume less political media",
    "Apathetic/protest non-voters are less socially integrated"
  ),
  Supported = c(
    "Supported",
    "Partially supported",
    "Supported",
    "Supported",
    "Not supported",
    "Not supported"
  )
)

# Render table
kable(
  hypotheses_clean,
  caption = "Table X: Summary of Hypotheses and Empirical Support",
  col.names = c("Hypothesis", "Description", "Empirical Support"),
  align = c("c", "l", "c")
)

```

In summary, **Models 1 and 2 provide strong and consistent evidence** for a motivational and socioeconomic explanation of persistent non-voting. **Model 2 performs best**, both in terms of statistical fit and theoretical clarity. Model 3, while methodologically improved through grouped predictors, ultimately reinforces rather than extends the core findings.

# Discussion

The results of this study offer clear empirical support for a typology-based understanding of persistent abstention in Switzerland. Most notably, the first two models demonstrate that motivational and structural factors significantly shape patterns of electoral participation. Political interest stands out as the most consistent predictor, confirming the theoretical expectations derived from the Civic Voluntarism Model (CVM). The strong effects in Model 1 highlight that apathetic non-voters are significantly less politically engaged than voters or other abstainer types.

Similarly, Model 2 reveals that lower levels of education and income are strongly associated with apathy-driven abstention. This confirms the notion that resource-based inequalities—both cognitive and material—create barriers to political participation. Satisfied non-voters, by contrast, closely resemble voters in terms of socioeconomic characteristics, which supports the idea of rational disengagement rather than marginalization.

The protest-oriented group, while conceptually distinct, remains more elusive empirically. Although Model 1 provides partial support for Hypothesis 2, institutional trust appears less strongly predictive than expected. This may reflect the relatively small group size of protest abstainers, or suggest that protest abstention is more context-sensitive and thus harder to detect through attitudinal variables alone.

Model 3 aimed to test whether communicative and integrative variables (e.g. media consumption, political discussion) add explanatory power. However, even after simplifying the categorical predictors to avoid convergence issues, none of the additional variables reach statistical significance. This suggests that abstention in Switzerland is not primarily driven by information deficits or social isolation, but by deeper motivational and structural cleavages.

Importantly, these findings underscore that non-voters are not simply less informed or less connected—but rather differentially resourced and motivated. As such, normative concerns about political exclusion should focus primarily on the apathetic group, which combines low interest with low structural resources. By contrast, satisfied abstainers may not be politically marginalized, but instead selectively disengaged for rational reasons.

# Limitations 

This study has several limitations that should be considered when interpreting the results.

First, the **small number of cases** in the protest ($n = 43$) and satisfied non-voter ($n = 17$) groups limits statistical power and increases uncertainty in the estimates. Findings for these categories, especially the latter, should be viewed with caution.

Second, the classification of non-voter types is based on **self-reported reasons for abstention**, which may be affected by **social desirability bias** or post-hoc rationalization. Theoretical categories may oversimplify more complex individual motivations.

Third, to ensure model stability, several independent variables were **simplified or grouped**, potentially reducing nuance and masking relevant subgroup differences.

Lastly, the analysis is based on **cross-sectional panel data**, which does not allow for causal inference or capture changes over time.

Despite these limitations, the study offers important insights into the structure of persistent non-voting in Switzerland and provides a solid basis for further research.

# Conclusion

This paper examined the determinants of persistent non-voting in Switzerland using data from the SELECTS 2019–2023 panel study. Building on existing typologies, it classified abstainers into three groups—apathetic, protest-oriented, and satisfied—and tested a series of theory-driven hypotheses using multinomial logistic regression.

The analysis shows that political motivation and socioeconomic status are the most robust predictors of abstention type. Apathetic abstainers are systematically less interested in politics and face structural barriers to participation. Protest abstainers differ less consistently from voters, while satisfied abstainers resemble voters in most respects but choose not to participate for strategic or normative reasons.

The study further demonstrates that media use and social integration, while often emphasized in the literature, do not appear to meaningfully explain persistent abstention in the Swiss case. This suggests that outreach efforts aimed at increasing participation should focus more on improving political efficacy and addressing structural inequalities than on purely communicative interventions.

In sum, this research highlights the importance of distinguishing between different types of non-voters. While some forms of abstention may be normatively benign, others reflect deeper patterns of exclusion. Differentiated engagement strategies—tailored to the needs and constraints of each group—are therefore essential for strengthening democratic participation in Switzerland.

\newpage
# References
