---
title: "Non-voters in Switzerland"
subtitle: "A Data Analysis non voting"
author: 
  - "Felix Sigrist, felix.sigrist@stud.unilu.ch.com, 19-641-968"
  - "Supervisor: Prof. Dr. Benjamin Schlegel"
  - "Number of words: 2200"
lang: en
date: today
date-format: "D. MMMM YYYY"
format: 
  pdf:
    documentclass: scrreprt
    number-sections: true
    toc: true
    toc-depth: 2
    lof: false
    lot: false
    include-in-header:
      - text: |
          \usepackage{authblk}
          \usepackage{titling}
          \RedeclareSectionCommand[beforeskip=5pt plus 1pt minus 1pt]{section}
          \RedeclareSectionCommand[beforeskip=10pt plus 2pt minus 2pt]{chapter}
fontsize: 12pt
linestretch: 1.5
geometry: left=2cm,right=2cm,top=2.5cm,bottom=2.5cm
bibliography: nonvoting.bib
citeproc: true
execute:
  echo: false
  warning: false
knitr:
  opts_chunk:
    message: false
---

# Abstract {.unnumbered}

This paper examines how different types of non-voters in Switzerland can be classified and to which degree this classification models can predict whether someone votes or does not vote.

# Introduction

Although elections are a cornerstone of democratic legitimacy, more than half of the Swiss electorate regularly abstains from voting in federal elections. With voter turnout consistently below 50 percent since the 1970s, electoral abstention is not a a marginal anomaly but a structural feature of Swiss democracy[@freitag_labor_2005]. This disengagement raises many questions about democratic representation and political inclusion. 

Despite the scale of this phenomenon, the empirical study of it has long been underdeveloped in Switzerland. As Bühlmann, Freitag and Vatter [@vatter_pdf_2003] observed, Swiss electoral research seems to be showing an "astonishing restraint" in adressing non-voting empirically - even though Switzerland, toghether with the Unied States, ranks among the countries with the lowest voter turnout in national elections accross established democracies. This international positioning is often noted in comparative electoral studies [@freitag_labor_2005; @ostrander_using_2021].

In recent years, however, the traditional assumption that abstention equals apathy has been more and more challenged. Typological and theory-driven approaches

```{r}
library(ggplot2)

#read in data
Participation<-read.csv("Daten/NRW-Zeitreihe.csv",sep = ";")

#filter for national data
Participation_national<- subset(Participation, geoLevelName == "Schweiz")

ggplot(Participation_national, aes(x = wahl_jahr, y = wahlbeteiligung)) +
  geom_line(color = "#0072B2", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Voter Turnout in Swiss National Council Elections",
    x = "Election Year",
    y = "Turnout (%)"
  ) +
  theme_minimal(base_size = 13) +
  scale_x_continuous(breaks = seq(min(Participation_national$wahl_jahr), max(Participation_national$wahl_jahr), by = 4)) +
  ylim(30, 60)



#| echo: false
#| eval: false
#| error: false
#| label: 

```

# Literature Review

Political non-voting is a central theme in research on Swiss democracy. Several studies have aimed to typologize the heterogeneous group of non-voters and explain their behavior using theoretical models and empirical typologies.

## Classical Approaches to Political Participation

Earlier research on voter turnout was largely grounded in the **socio-structural** approach (e.g., Lazarsfeld), the **social-psychological** model (e.g., Campbell et al.), and **rational choice** theory (e.g., Downs). These were later extended through the **Standard Model of Political Participation** (Barnes & Kaase, 1979), which emphasized the role of socioeconomic resources and cognitive mobilization [@vatter_pdf_2003].

A conceptual advancement came with the introduction of the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which identifies three central factors influencing political participation: **resources**, **motivation**, and **recruitment networks**. The CVM has proven particularly useful in explaining unequal patterns of participation and in identifying different types of abstainers.

## Milbrath’s Hierarchical Model of Participation

Milbrath’s framework [@ruedin_testing_2007] conceptualizes political participation as a **cumulative hierarchy**—ranging from passive forms of involvement (e.g., voting, discussing politics) to more active and elite-level activities (e.g., running for office). This model combines individual characteristics with exposure to political stimuli and has been validated through simulation studies that demonstrate the dynamic interplay of individual and institutional factors.

## A Swiss Typology: Bühlmann et al. (2003)

[@vatter_pdf_2003] developed one of the first typologies of Swiss non-voters using cluster analysis on survey data from the 1999 federal elections. Their typology includes six distinct types:

-   Politically Disinterested\
-   Politically Disenchanted\
-   Socially Isolated\
-   Politically Incompetent\
-   Protest Voters\
-   Alternative Participants

These types differ in terms of political interest, efficacy, trust in institutions, and socioeconomic resources.

## Extension and Replication: Fatke & Freitag (2015)

[@fatke_wollen_2015] built upon this typology using SELECTS 2011 data. Their cluster analysis refined the groupings and reaffirmed the heterogeneity of non-voters. Types such as *satisfied but disengaged*, *incompetent*, and *abstaining but participatory* highlight that abstention does not always stem from apathy or alienation, but also from rational disengagement or alternative political expression.

## Democratic Implications and Empirical Consensus

Across all studies, a key insight prevails: **non-voters are not a homogeneous group**. From a democratic theory perspective, the most problematic are the **socially isolated** and **incompetent**, who lack both the resources and social networks for political engagement [@fatke_nichtwahlenden_2015].

At the same time, not all abstention is a cause for concern. **Alternative participants** or the **satisfied disengaged**, for example, abstain not due to rejection or apathy, but because they either prefer other forms of participation (e.g., initiatives, petitions) or are fundamentally satisfied with the system.

This differentiated understanding provides the conceptual bridge to the following theory section, which formalizes these empirical distinctions into a testable framework.

------------------------------------------------------------------------

# Theory

## Electoral Abstention as a Multi-Dimensional Phenomenon

Political abstention is increasingly seen as a multidimensional phenomenon. Contemporary research emphasizes that non-voters differ systematically in terms of political motivation, cognitive engagement, social integration, and structural access. Typological approaches have shown that abstention reflects distinct profiles of political disengagement, rather than a binary act of participation versus non-participation [@vatter_pdf_2003; @fatke_wollen_2015].

## The Civic Voluntarism Model as Theoretical Foundation

This study applies the **Civic Voluntarism Model (CVM)** [@ostrander_using_2021], which explains political engagement through three core components:

1.  **Resources** – time, money, civic skills, and knowledge\
2.  **Motivation** – interest in politics, efficacy, ideological conviction\
3.  **Recruitment Networks** – social integration, group membership, mobilization

According to CVM, individuals are more likely to participate politically when they are motivated, able, and socially embedded. Conversely, the absence of any one of these factors can lead to abstention. Crucially, access to these dimensions is unequally distributed—meaning that political inequality is reproduced through structural exclusion [@ostrander_using_2021].

The CVM therefore provides a robust foundation for explaining **persistent abstention**, as it integrates individual capacities, perceptions, and social context into one explanatory model.

## Typologies of Non-Voters in Switzerland

Building on the CVM, the typologies developed by Bühlmann et al. [@vatter_pdf_2003] and extended by Fatke & Freitag [@fatke_wollen_2015] offer a detailed framework for classifying non-voters. For the purpose of this study, these types are aggregated into three theoretically meaningful categories:

| Type | Main Mechanism | Theoretical Foundation |
|----------------------|-------------------------|-------------------------|
| **Apathetic** | Lack of motivation/resources | Civic Voluntarism Model |
| **Protest-oriented** | Political alienation | System support/trust theory |
| **Satisfied/Alternative** | Rational disengagement | Expressive/instrumental logic |

This typology provides the conceptual basis for analyzing **persistent non-voters**, i.e., those who abstained in both 2019 and 2023 federal elections in Switzerland.

## Persistent Non-Voting as an Analytical Focus

Persistent abstention is of particular interest, as it suggests more entrenched and stable forms of disengagement. Unlike situational abstention, which may result from temporary constraints, persistent non-voters are likely to exhibit deeper structural or attitudinal barriers. Typologizing this group allows for a theory-driven understanding of *why* certain individuals consistently opt out of electoral politics.

## Theoretical Expectations and Hypotheses

Based on the typology and the Civic Voluntarism Model, the following hypotheses are derived:

-   **H1 (Political Interest & Knowledge):** Apathetic non-voters report significantly lower political interest and political knowledge than other groups.

-   **H2 (System Satisfaction & Trust):** Protest-oriented non-voters express lower trust in political institutions and lower satisfaction with democracy than other groups.

-   **H3 (Socioeconomic Status):** Apathetic non-voters have lower levels of education and income, while satisfied/alternative non-voters resemble voters in their socioeconomic status.

-   **H4 (Social Integration):** Apathetic non-voters show weaker social integration (e.g., fewer memberships or interpersonal networks) than other groups.

-   **H5 (Party Identification & Ideological Self-Placement):** Protest-oriented non-voters are ideologically positioned but report weak or no party identification.

These hypotheses will be empirically tested using a multinomial logistic regression model, treating **voters** as the reference category. The predictors reflect the theoretical dimensions discussed above.

# Empirical Analysis

## Data and Descriptive Statistics

We use data from the CSES dataset, focusing on economic evaluations, incumbent support, and controls. Descriptive statistics are summarized in @tbl-stats.

```{r}

```

## Methods

We estimate an OLS model and a logistic regression with interaction terms. The dependent variable is voting for the incumbent, and key predictors include economic evaluations and political information.

## Results

OLS results are shown in @tbl-reg.

Logistic regression results explore interaction effects, visualized in @fig-interaction.

# Discussion

The results confirm that negative economic perceptions reduce incumbent support. Politically informed voters show a stronger reaction, highlighting the moderating role of knowledge.

# Conclusion

This study supports economic voting theory, emphasizing political information as a critical moderator. Future research should explore additional factors influencing voting behavior.
